\cleardoublepage
\thispagestyle{plain}

\section*{Acknowledgments}


I am deeply grateful to my principal supervisor, Prof. Dr. Martin Leucker, for providing the opportunity and necessary conditions to conduct this dissertation. He gave me enough freedom and trust to pursue my research independently, while also maintaining the right amount of supervision to keep the work focused. Working with Martin not only increased the number of contributions but also made them more impactful and formally sound. I have learned many scientific skills during our brief consultations, and now that I have finished, I wish we had more time for discussion.

Special thanks are also due to Prof. Dr. Gordon J. Pace,  the second jury member, for taking the time to review my dissertation and traveling all the way from Malta to attend my defense. I highly appreciated his precise, meticulous, and constructive criticism of my work and his enthusiastic view of its connection to other works and its potential continuation.

I extend my thanks to Prof. Dr. Stefan Fischer, chair of the examination committee. His professional and efficient management ensured a seamless atmosphere for the defense and and made the subsequent administrative processes straight forward.

I extend my gratitude to my second supervisor, Prof. Dr. Gerardo Schneider, for introducing me to the field of normative reasoning.  He influenced this work by his guidance on the necessary conceptual shift from traditional software verification to the subtleties of legal and normative reasoning. The literature and the specific research problems he provided served as the starting point for this dissertation.

I am also very grateful to Shaun Azzopardi for the work we did together, which contributed to the second Part of this dissertation. Our collaboration and teamwork began during my visit to Gothenburg University in August 2021 and lasted for two years, through a series of rejections. He kept working with me and remained enthusiastic until our work was published in 2023.

Deep appreciation is owed to my colleague and office neighbor, Daniel Thomas, for his continuous scientific and moral support. Having him by my side during these challenging years was a pleasant and insightful experience. While I regret not having worked together with him on a publication. I keep hoping to get the opportunity to collaborate on a project in the near future.

A cornerstone of this dissertation's success is the support from my family and friends. My mother and two sisters are my  pillars of inconditional love and biggest source of motivation, who, alongside my friends Anouar, Ahmed, and a certain Velba (an angel I believe was sent from the seventh sky), have shown me their abundant support, especially during the darkest and toughest moments of this journey.

Finally, I would like to thank my former colleagues at the institute since 2018 for their sincere camaraderie. These ``specimens'' of kind research staff accompanied me with forward-thinking discussions and valuable networking opportunities. I look forward to remaining an active and contributing member of their network.