\subsection{Monitor Construction for Tight Contract Satisfaction}

In the previous section, we established the mathematical rules that define exactly when a contract is satisfied or violated. 
Now, we turn those abstract definitions into an operational implementation: a finite-state monitor that tracks compliance step-by-step as events occur. 
We design these as Moore-style machines, which means the monitor maintains a running summary of the interaction; every state corresponds directly to one of the five possible verdicts (such as "tight satisfaction" or "undecided"). By ensuring these monitors update deterministically with every new event, we create a reliable, automated way to check compliance that matches our theoretical rules perfectly.
\begin{definition}[Moore machine on tight satisfaction verdicts]
    \label{def:tightsatmon}
    The \emph{Moore machine on tight satisfaction verdicts}, written $\tmon$, is a Moore machine whose output alphabet is the five-valued verdict set $\tightverdicts$.
    That is:
    \[
    \tmon = (Q,q_0,\Gamma,\tightverdicts,\delta,\lambda_{5}),
    \]
    where:
    \begin{enumerate}
    \item The output alphabet formed by 5 letters is 
    \[
    \tightverdicts = \{\mathsf{?},\topt,\bott,\topp,\botp\}.
    \]
    \item $Q$ is the set of states and $q_0\in Q$ is the initial state,
    \item $\Gamma$ is the input event alphabet,
    \item $\delta: Q \times \Gamma \to Q$ is the transition function,
    \item $\lambda_{5}: Q \to \tightverdicts$ is the state output function.
    \end{enumerate}
    \end{definition}

The next definition introduces the construction for any contract into its tight
satisfaction monitor. The construction is a function that maps each contract $C$ to a tight satisfaction monitor that enforces it. The construction proceeds by structural induction on the syntax of $C$, and each operator in \cDL is matched by a corresponding monitor combination operator. Regular expression guards and triggers rely on the
tight satisfaction monitor $\tsmc(\re)$ introduced earlier in
Definition~\ref{def:tsmc-re}.

\begin{definition}[Tight Satisfaction Monitor Construction]
 \label{def:tsmc}
The \emph{tight satisfaction monitor construction} is a function defined on $\cDL$, written  $\tsmc(C)$, that returns the tight satisfaction monitor for a contract
$C$ in \cDL. It is defined inductively on the structure of $C$:
 \[
\tsmc(C) \;:=\;
\begin{cases}
  \tsmc_{\mathit{lit}}(\ell)
    & \text{if } C = \ell, \\[0.4em]

  \tsmc_{\mathit{\wedge}}\big(\tsmc(C_1),\tsmc(C_2)\big)
    & \text{if } C = C_1 \wedge C_2, \\[0.4em]

  \tsmc_{\mathit{;}}\big(\tsmc(C_1),\tsmc(C_2)\big)
    & \text{if } C = C_1 ; C_2, \\[0.4em]

  \tsmc_{\mathit{\repair}}\big(\tsmc(C_1),\tsmc(C_2)\big)
    & \text{if } C = C_1 \repair C_2, \\[0.4em]

  \tsmc_{\trigg}(\re,C')
    & \text{if } C = \trig[\re]{C'}, \\[0.4em]

  \tsmc_{\guardd}(\re,C')
    & \text{if } C = \guard[\re]{C'}, \\[0.4em]

  \tsmc_{\mathit{nrep}}\big(n,\tsmc(C')\big)
    & \text{if } C = (C')^n, \\[0.4em]

  \tsmc_{\mathit{Rep}}\big(\tsmc(C')\big)
    & \text{if } C = \repit{C'}\;.
\end{cases}
\]
Where the tight monitor construction for regular expressions $\tsmc(\re)$ is already defined in Definition~\ref{def:tsmc-re}.
\end{definition}

\paragraph{Monitor correctness invariant.}
The monitor construction preserves the following invariant.
For every contract $C$ in \cDL, every finite trace $\pi$, and every prefix $\pi_k$,
the state reached by the monitor $\tsmc(C)$ after reading $\pi_k$
emits exactly the verdict prescribed by the five-valued tight semantics, that is:
\[
\lambda_5\bigl(\delta(q_0,\pi_k)\bigr) \;=\; \semfive{\pi_k \vDash C}.
\]
All operator-specific constructions are designed to preserve this invariant under
synchronous product, redirection, and state elimination.
Correctness lemmas below establish that the invariant holds inductively for each
contract constructor. 

We construct in the next subsections the tight satisfaction monitors compositionally from the syntax of contracts.
The construction proceeds by structural induction on the contract and associates to each
contract operator a corresponding monitor-combination operator.
Conceptually, this follows the classical construction paradigm of Thompson for regular expressions,
where complex behaviors are obtained by composing simpler automata.
However, rather than constructing an intermediate nondeterministic automaton and determinizing it afterward,
we perform the construction directly at the level of deterministic Moore machines.
In particular, verdict propagation is built into the states and outputs of the monitor from the outset,
so that tight satisfaction and violation are observed incrementally on prefixes of the trace.
This avoids a separate determinization step and ensures that the monitor semantics coincides
by construction with the five-valued tight semantics.

\subsubsection*{Construction for Literal Contracts}
\paragraph{From Tight Semantics to 5-Valued Monitoring.}
The literal clauses above define one-step satisfaction and violation judgements for a single event word $\trace{A}$.
We lift these clauses into a five-valued Moore machine whose outputs track the evolution of the tight verdicts over prefixes.
\begin{definition}[Tight Satisfaction Monitor Construction for Literals]
\label{def:moore-5-literal}
For a literal $\ell$ from \cDL, the \emph{Tight Satisfaction monitor construction} for $\ell$, written $\tsmc_{\mathit{lit}}(\ell)$ is defined as:
\[
\tsmc_{\mathit{lit}}(\ell)
  = (Q, q_0, \Gamma, \tightverdicts, \delta, \lambda_5).
\]
\begin{itemize}
  \item $Q=\{q_0,q_s,q_v,q_{ps},q_{pv}\}$,
        with outputs
        $\lambda_5(q_0)=\mathsf{?}$,
        $\lambda_5(q_s)=\topt$,
        $\lambda_5(q_v)=\bott$,\\
        $\lambda_5(q_{ps})=\topp$,
        $\lambda_5(q_{pv})=\botp$.
  \item $\Gamma=2^{\Sigma}$ is the event alphabet.
  \item $\delta:Q\times\Gamma\to Q$ is defined as follows:
\[
\begin{aligned}
&\text{(1) tight transition: } 
&& \delta(q_0,A)=
  \begin{cases}
    q_s & \text{if }\trace{A}\ \satt\ \ell,\\
    q_v & \text{if }\trace{A}\ \violt\ \ell;
  \end{cases}\\[2pt]
&\text{(2) post transitions: }
&& \delta(q_s,A)=q_{ps},\ \delta(q_v,A)=q_{pv},\ \\ & &&
   \delta(q_{ps},A)=q_{ps},\ \delta(q_{pv},A)=q_{pv}.
\end{aligned}
\]

\end{itemize}
\end{definition}

\paragraph{Intuition.} For every atomic literal, the monitor construction initializes at $q_0$, emitting the undecided verdict $\mathsf{?}$ to reflect the absence of decisive information.
Upon consuming the first event, the execution crosses the decisive frontier, transitioning to either $q_s$ (success at frontier) with verdict $\topt$ if the literal is satisfied, or $q_v$ (violation at frontier) with verdict $\bott$ if it is violated.
On the next step after this decision, the monitor enters the stable regions $q_{ps}$ (post-success) or $q_{pv}$ (post-violation), where it permanently emits $\topp$ or $\botp$ respectively.
These indices explicitly encode the temporal relationship to the frontier, distinguishing the exact moment of decision from the immutable history of the trace.

\begin{figure}[h!]
\centering
\begin{tikzpicture}[
shorten >=1pt,auto,node distance=28mm,semithick,
  every state/.style={rectangle,rounded corners,draw,minimum width=11mm,
    minimum height=7.5mm,inner sep=5pt,font=\footnotesize,align=center}
]

% --- states ---------------------------------------------------
\node[initial,state,fill=gray!10] (q0)  {$q_0$\\[0.2pt]$\mathsf{?}$};
\node[state,fill=green!18,right=of q0] (qs)  {$q_s$\\[0.2pt]$\topt$};
\node[state,fill=red!18,below=of q0]   (qv)  {$q_v$\\[0.2pt]$\bott$};
\node[state,fill=green!10,right=of qs] (qps) {$q_{ps}$\\[0.2pt]$\topp$};
\node[state,fill=red!10,below=of qps]  (qpv) {$q_{pv}$\\[0.2pt]$\botp$};

% --- transitions ----------------------------------------------
\path[->]
  (q0) edge[bend left=12] node[above,pos=0.5]
    {$\{a^{(1)},a^{(2)}\}\subseteq A$} (qs)
  (q0) edge[bend right=12] node[left,pos=0.45]
    {$\{a^{(1)},a^{(2)}\} \not\subseteq A$} (qv)
  (qs)  edge node[above] {$\Gamma$} (qps)
  (qv)  edge node[below] {$\Gamma$} (qpv)
  (qps) edge[loop right] node {$\Gamma$} (qps)
  (qpv) edge[loop right] node {$\Gamma$} (qpv);
\end{tikzpicture}
\caption{Compact 5-verdict Moore machine for the obligation literal
$\obl[1]{a}$. 
Each node displays its internal state and the corresponding output verdict $\in\tightverdicts$.
The first joint execution of $a^{(1)}$ and $a^{(2)}$ yields~$\topt$,
otherwise~$\bott$; subsequent steps emit the post-frontier verdicts
$\topp$ or $\botp$.}
\label{fig:moore-obligation-compact}
\end{figure}


\subsubsection*{Construction for Binary Contract Operators}

For binary contract operators of the form $C\ \textsf{op}\ C'$ with
$\textsf{op} \in \{\wedge,\ ;\ ,\ \repair\}$, the monitor for the composite contract is obtained by combining the already constructed monitors
$\tsmc(C)$ and $\tsmc(C')$. Each operator has its own monitor construction,
defined below for conjunction, sequence, and reparation. We begin with the
conjunction case.


\begin{definition}[Tight Satisfaction Monitor Construction for Conjunction]
\label{def:moore-5-conj}
Let $C$ and $C'$ be contracts in \cDL, and let
\[
\tsmc(C)  = (Q, q_0, \Gamma, \tightverdicts, \delta, \lambda_5)
\quad\text{and}\quad
\tsmc(C') = (Q', q'_0, \Gamma, \tightverdicts, \delta', \lambda'_5).
\]
The tight satisfaction monitor construction for the conjunction
$C \wedge C'$, written $\tsmc_{\wedge}(C,C')$, is defined as:
\[
\tsmc_{\wedge}(C,C')
  = (Q_{\wedge}, q_0^{\wedge}, \Gamma, \tightverdicts, \delta_{\wedge}, \lambda_5^{\wedge}).
\]

\begin{itemize}
  \item The state set is the Cartesian product
  \[
    Q_{\wedge} = Q \times Q',
  \]
  with the initial state
  \[
    q_0^{\wedge} = (q_0,\,q'_0).
  \]

  \item The output function is
  \[
    \lambda_5^{\wedge}(x,y) =
    \lambda_{5}^{\textsf{comb}}\!\big(\lambda_5(x),\ \lambda'_5(y)\big),
  \]
  where $\lambda_{5}^{\textsf{comb}}$ is the conjunction-combination table:
  \[
  \begin{array}{c|ccccc}
  \lambda_{5}^{\textsf{comb}}(v_1,v_2) & \mathsf{?} & \topt & \bott & \topp & \botp\\\hline
  \mathsf{?} & \mathsf{?} & \mathsf{?} & \bott & \mathsf{?} & \botp\\
  \topt      & \mathsf{?} & \topt      & \bott & \topt      & \botp\\
  \bott      & \bott      & \bott      & \bott & \bott      & \botp\\
  \topp      & \mathsf{?} & \topt      & \bott & \topp      & \botp\\
  \botp      & \botp      & \botp      & \botp & \botp      & \botp
  \end{array}
  \]

  \item The transition function is the synchronous product:
  \[
    \delta_{\wedge}\big((x,y),A\big)
      = \big(\delta(x,A),\ \delta'(y,A)\big),
  \]
  for all $(x,y)\in Q_{\wedge}$ and $A\in\Gamma$.
\end{itemize}
\end{definition}


\noindent
\textbf{Partial dominance rules for $\wedge$.}
The monitor for $C \wedge C'$ checks both components at the same time and decides the global verdict according to the following rules:

\begin{itemize}
  \item If any component gives $\botp$, the result is $\botp$:
  \[
  \forall v\in\tightverdicts:\quad \botp \sqcap v = \botp,
  \]
  That is, once a permanent violation appears, the whole conjunction is permanently violated.
  \item If any component gives $\bott$, and none is permanent, the result is $\bott$:
  \[
  \forall v\in\{\mathsf{?},\topt,\topp\}:\quad \bott \sqcap v = \bott,
  \]
  That is, a single tight violation makes the conjunction fail tightly.
  \item Tight and permanent success combine as the weakest success:
  \[
  \topt \sqcap \topt = \topt,\qquad
  \topt \sqcap \topp = \topt,\qquad
  \topp \sqcap \topp = \topp,
  \]
  The conjunction is only permanently satisfied when both parts are permanent.
  \emph{Note:} The case $\topt \sqcap \topp = \topt$ reflects that if one component is exactly at its satisfaction frontier ($\topt$) while the other is already past it ($\topp$), the conjunction as a whole is determined by the later of the two, effectively placing the global frontier at the current step.
  \item If both sides are undecided or only partly satisfied, the result stays $\mathsf{?}$:
  \[
  \mathsf{?}\sqcap v = \mathsf{?}\quad\text{for }v\in\{\mathsf{?},\topt,\topp\},
  \]
  The monitor waits until a clear outcome appears.
  \item The operator is symmetric:
  \[
  v_1\sqcap v_2 = v_2\sqcap v_1,
  \]
  The order of operands does not matter.
\end{itemize}


\begin{lemma}[Correctness of the conjunctive monitor construction]
\label{lem:conj-correct}
Let 
\[
\tsmc(C)  = (Q, q_0, \Gamma, \tightverdicts, \delta, \lambda_5)
\quad\text{and}\quad
\tsmc(C') = (Q', q'_0, \Gamma, \tightverdicts, \delta', \lambda'_5)
\]
and let
\[
\tsmc(C \wedge C')
  = (Q_{\wedge}, q_0^{\wedge}, \Gamma, \tightverdicts, \delta_{\wedge}, \lambda_5^{\wedge})
\]
be the conjunction monitor defined in Definition~\ref{def:moore-5-conj}.
For every trace $\pi$, the output of the conjunction monitor satisfies:
\[
\lambda_5^{\wedge}(\delta_\wedge(q_0^\wedge,\pi))=
\semfive{\pi \vDash C \wedge C'}.\]
\end{lemma}

\noindent
\textbf{Proof sketch.}
The monitor $\tsmc_{\wedge}(\tsmc(C),\tsmc(C'))$ runs both component monitors in parallel and
computes its output using the conjunction-combination table
$\lambda_{5}^{\textsf{comb}}$. The correctness follows from the prefix-based
tight semantics of $C \wedge C'$.

\begin{itemize}
  \item Permanent violation in either component produces $\botp$ immediately.
  \item A tight violation in one component produces $\bott$ whenever no permanent result is already present.
  \item Satisfaction requires both components to reach satisfaction states.  
        If one is in $\topp$ or $\topt$ and the other is non-violating, the
        combined verdict matches the corresponding entry in the table.
  \item If both components remain undecided, the output is $\mathsf{?}$.
\end{itemize}

These cases match exactly the clauses for tight satisfaction, tight violation,
post-satisfaction, and post-violation for the contract $C \wedge C'$. The monitor, therefore, correctly implements the tight semantics of conjunction.
\qed


Sequential composition is the second binary operator of \cDL. Given two already
constructed monitors $\tsmc(C)$ and $\tsmc(C')$, the monitor for the composite
contract $C\,;\,C'$ must first execute $C$ on the incoming trace and, once $C$
reaches tight satisfaction, must continue execution with $C'$ on the remaining
suffix. The construction below reuses the state spaces of both components by redirecting transitions that correspond to the tight success of $C$
into the initial state of $C'$. This yields a tight prefix monitor that exactly matches the semantics of sequential composition.

\begin{definition}[Sequential Monitor Construction]
\label{def:moore-seq}
Let
\[
\tsmc(C)  = (Q, q_0, \Gamma, \tightverdicts, \delta, \lambda_5)
\quad\text{and}\quad
\tsmc(C') = (Q', q'_0, \Gamma, \tightverdicts, \delta', \lambda'_5)
\]
be the tight satisfaction monitors of $C$ and $C'$.  
The tight satisfaction monitor for the sequential composition
$C\,;\,C'$, written $\tsmc_{\bm{;}}(C,C')$, is defined as:
\[
\tsmc_{\bm{;}}(\tsmc(C),\tsmc(C'))
  = (Q_{;}, q_0^{;}, \Gamma, \tightverdicts, \delta_{;}, \lambda_5^{;}).
\]

\begin{itemize}
  \item The state set is
  \[
    Q_{;} = (Q \setminus Q^{+})\ \cup\ Q',
  \]
  where
  \[
    Q^{+} = \{\,x \in Q \mid \lambda_5(x) \in \{\topt,\topp\}\,\},
  \]
  and the initial state is
  \[
    q_0^{;} = q_0.
  \]

  \item The transition function is
  \[
  \delta_{;}(q,A) =
  \begin{cases}
    q'_0 
      & \text{if } q\in Q \text{ and } \lambda_5(\delta(q,A)) = \topt, \\[4pt]
    \delta(q,A)
      & \text{if } q\in Q \text{ and } \lambda_5(\delta(q,A)) \notin \{\topt,\topp\}, \\[4pt]
    \delta'(q,A)
      & \text{if } q\in Q'. \\
  \end{cases}
  \]

  \item The output function is
  \[
  \lambda_5^{;}(x) =
  \begin{cases}
    \lambda_5(x)  & \text{if } x\in Q,\\
    \lambda'_5(x) & \text{if } x\in Q'. \\
  \end{cases}
  \]
\end{itemize}
\end{definition}

\noindent
\textbf{Intuition.}
The construction implements the idea that $C$ must succeed tightly before
$C'$ becomes active. All states of $C$ that already correspond to tight
satisfaction ($\lambda_5(x)=\topt$ or $\topp$) are removed, since execution
should never continue inside them. Any transition in $C$ that would have
entered such a removed state is redirected to the initial state $q'_0$ of
$C'$, thereby starting the second contract at the exact prefix where $C$
achieves tight success. All other transitions behave exactly as in the
original monitors. The resulting machine therefore behaves as $C$ until $C$
succeeds tightly, after which it behaves as $C'$ for the remainder of the
trace.

\begin{lemma}[Correctness of the sequential monitor construction]
\label{lem:seq-correct}
Let $\tsmc(C)$ and $\tsmc(C')$ be the monitors for $C$ and $C'$,
and let $\tsmc_{\bm{;}}(\tsmc(C),\tsmc(C'))
= (Q_{;}, q_0^{;}, \Gamma, \tightverdicts, \delta_{;}, \lambda_5^{;})$. For every trace $\pi$, the monitor outputs the same verdict according to the sequence semantics:
\[
\lambda_5^{;}\big(\delta_{;}(q_0^{;},\pi)\big)=
\semfive{\pi \vDash C;C'}.
\]
\end{lemma}
\textbf{Proof.}
The proof is by induction on the length of the input trace $\pi$.

    \begin{proof}
        Let $M:=\tsmc_{\bm{;}}(\tsmc(C),\tsmc(C'))=(Q_{;}, q_0^{;}, \Gamma, \tightverdicts, \delta_{;}, \lambda_5^{;})$.
        We prove by induction on $n:=|\pi|$ that
        \[
        \lambda_5^{;}\big(\delta_{;}(q_0^{;},\pi)\big)
        \;=\;
        \semfive{\pi \vDash C;C'}.
        \]
        
        \paragraph{Base case ($n=0$).}
        For $\pi=\emptytrace$ we have $\delta_{;}(q_0^{;},\emptytrace)=q_0$ and thus
        \[
        \lambda_5^{;}\big(\delta_{;}(q_0^{;},\emptytrace)\big)=\lambda_5(q_0)=\mathsf{?}.
        \]
        This matches the tight semantics of $C;C'$ on the empty trace: no decisive prefix for $C$ has been reached and $C'$ cannot be active yet.
        
        \paragraph{Inductive step.}
        Assume the claim holds for all traces of length $n$.
        Let $\pi$ be a trace of length $n{+}1$ and write $\pi=\pi'\concat\trace{A}$ with $|\pi'|=n$ and $A\in\Gamma$.
        Let $s:=\delta_{;}(q_0^{;},\pi')$ be the state reached after reading $\pi'$.
        We distinguish whether the monitor is already executing the right component.
        
        \smallskip
        \noindent\emph{Case 1: $s\in Q'$.}
        Then, by Definition~\ref{def:moore-seq}, $\delta_{;}(s,A)=\delta'(s,A)$ and $\lambda_5^{;}(s)=\lambda'_5(s)$, hence
        \[
        \lambda_5^{;}\big(\delta_{;}(q_0^{;},\pi)\big)
        =
        \lambda'_5\big(\delta'(s,A)\big).
        \]
        Reaching a state of $Q'$ means that the switch from $C$ to $C'$ has already happened at some earlier prefix, and from that point on the sequential monitor behaves exactly as the monitor for $C'$ on the remaining suffix.
        Therefore the tight semantics of $C;C'$ on $\pi$ is determined by the verdict of $C'$ on that suffix.
        By correctness of $\tsmc(C')$ (established for the tight monitor construction),
        the right-hand side equals $\semfive{\pi \vDash C;C'}$.
        
        \smallskip
        \noindent\emph{Case 2: $s\in Q$.}
        Let $t:=\delta(s,A)$ be the next state of the left monitor.
        There are two subcases.
        
        \smallskip
        \noindent\emph{Case 2(a): $\lambda_5(t)=\topt$.}
        Then Definition~\ref{def:moore-seq} redirects the transition to $q'_0$, so $\delta_{;}(s,A)=q'_0$ and
        \[
        \lambda_5^{;}\big(\delta_{;}(q_0^{;},\pi)\big)
        =
        \lambda_5^{;}(q'_0)
        =
        \lambda'_5(q'_0)
        =
        \mathsf{?}.
        \]
        Semantically, this is exactly the step where $C$ reaches its tight satisfaction frontier on the prefix $\pi$, and the contract $C'$ becomes active on the remaining suffix.
        At this switching moment, the suffix is empty, so the status of $C'$ is undecided, and the sequence verdict is $\mathsf{?}$.
        Thus the monitor output matches $\semfive{\pi \vDash C;C'}$.
        
        \smallskip
        \noindent\emph{Case 2(b): $\lambda_5(t)\notin\{\topt,\topp\}$.}
        Then the switch has not occurred, so $\delta_{;}(s,A)=t$ and $\lambda_5^{;}(t)=\lambda_5(t)$.
        Therefore
        \[
        \lambda_5^{;}\big(\delta_{;}(q_0^{;},\pi)\big)
        =
        \lambda_5(t)
        =
        \semfive{\pi \vDash C}.
        \]
        Before $C$ reaches tight satisfaction, the sequence semantics is governed by the left component on the current prefix, hence $\semfive{\pi \vDash C}=\semfive{\pi \vDash C;C'}$ in this case.
        So the equality holds.
        
        All cases agree with the semantic clauses of sequential composition, so the induction closes.
        \end{proof}



\begin{definition}[Reparation Monitor Construction]
\label{def:moore-repair}
Let
\[
\tsmc(C)  = (Q, q_0, \Gamma, \tightverdicts, \delta, \lambda_5)
\quad\text{and}\quad
\tsmc(C') = (Q', q'_0, \Gamma, \tightverdicts, \delta', \lambda'_5)
\]
be the tight satisfaction monitors for $C$ and $C'$.  
The monitor for the reparation contract $C\repair C'$, written
\[
\tsmc_{\repair}(C,C'),
\]
is defined as the tuple
\[
\tsmc_{\repair}(\tsmc(C),\tsmc(C'))
  = (Q_{\repair}, q_0^{\repair}, \Gamma, \tightverdicts, \delta_{\repair}, \lambda_5^{\repair}).
\]

\begin{itemize}
  \item The state set is
  \[
    Q_{\repair} = (Q \setminus Q^{-})\ \cup\ Q',
  \]
  where
  \[
    Q^{-} = \{\,q \in Q \mid \lambda_5(q) \in \{\bott,\botp\}\,\},
  \]
  and the initial state is
  \[
    q_0^{\repair} = q_0.
  \]

  \item The transition function is
  \[
    \delta_{\repair}(q,A)=
    \begin{cases}
      q'_0
        & \text{if } q\in Q \text{ and } \lambda_5(\delta(q,A))=\bott,\\[4pt]
      \delta(q,A)
        & \text{if } q\in Q \text{ and } \lambda_5(\delta(q,A))\notin\{\bott,\botp\},\\[4pt]
      \delta'(q,A)
        & \text{if } q\in Q'.\\
    \end{cases}
  \]

  \item The output function is
  \[
     \lambda_5^{\repair}(q)=
     \begin{cases}
       \lambda_5(q)   & \text{if } q\in Q,\\[2pt]
       \lambda'_5(q)  & \text{if } q\in Q'.\\
     \end{cases}
  \]
\end{itemize}
\end{definition}


\noindent
\textbf{Intuition.}
The reparation operator activates the secondary contract $C'$ after the primary contract $C$ reaches a tight failure. The construction mirrors the sequential case, except that the redirection applies to transitions leading to a tight violation of $C$ rather than to those leading to tight satisfaction.

All states of $C$ whose verdicts are already violating
($\lambda_5(q)\in\{\bott,\botp\}$) are removed. Every transition in $C$ that
would enter such a state is redirected to $q'_0$, the initial state of $C'$.
This starts $C'$ exactly at the tight prefix where $C$ tightly fails.  

As long as no violation occurs, the monitor behaves exactly as $C$. Once a tight failure is detected, the remaining input is processed by $C'$.

\begin{lemma}[Correctness of the reparation monitor construction]
\label{lem:rep-correct}
Let $\tsmc(C)$ and $\tsmc(C')$ be the monitors for $C$ and $C'$, and let
\[
\tsmc_{\repair}(\tsmc(C),\tsmc(C'))
  = (Q_{\repair}, q_0^{\repair}, \Gamma, \tightverdicts, \delta_{\repair}, \lambda_5^{\repair})
\]
be the monitor constructed in Definition~\ref{def:moore-repair}.
Then, for every trace $\pi$, the monitor outputs the right verdict as specified by the tight semantics:
\[
\lambda_5^{\repair}(\delta_{\repair}(q_0^{\repair},\pi)) = \semfive{\pi \vDash C \repair C'}.
\]
\end{lemma}
\begin{proof}[Proof sketch]
The argument follows exactly the same scheme as the proof of
Lemma~\ref{lem:seq-correct} for sequential composition.

The reparation monitor behaves as the tight monitor for $C$ as long as no tight violation occurs.
By the correctness invariant of $\tsmc(C)$, all prefixes before the first tight violation are evaluated exactly as prescribed by the tight semantics of $C\repair C'$.

At the unique prefix where $C$ reaches a tight violation ($\bott$), the construction redirects the transition to the initial state $q'_0$ of the monitor for $C'$.
This corresponds precisely to the semantic clause that activates the reparation contract after the first decisive failure of $C$.

From that point on, the monitor evolves exactly as $\tsmc(C')$ on the remaining suffix.
By the inductive correctness of the tight monitor construction for $C'$, the emitted verdict coincides with the tight semantics of $C'$ on that suffix.

Since all other transitions are unchanged and no ambiguity about the switching point is possible, the monitor output agrees with the tight semantics of $C\repair C'$ on all finite traces.
\end{proof}






\begin{remark}[Relation to Control Phases and Prior Work]
  \label{rem:control-phases}
  The structural constructions in Definitions~\ref{def:moore-seq} and~\ref{def:moore-repair} serve as the static counterpart to the \emph{control-phase} view described in runtime verification frameworks such as \cite{falcone2009runtime,bartocci2018rv}.
  While these standard frameworks often rely on an explicit control variable $m \in \{\mathbf{L}, \mathbf{R}\}$ to dynamically track whether the left ($\mathbf{L}$) or right ($\mathbf{R}$) component is active, our transformation embeds this logic directly into the transition graph.
  We achieve this by eliminating the terminal states of the left monitor and redirecting the decisive transitions—$\topt$ for sequence and $\bott$ for reparation—to the initial state of the right monitor.
  Consequently, this compile-time structure effectively encodes the operational behavior of a mode-augmented monitor that switches phases only when specific conditions are met.
  This design parallels the phase-based semantics where control modes synchronize sub-monitors for sequential patterns like ``\emph{after $C$ succeeds, check $C'$}''.
  Similarly, our reparation composition aligns with the ``\emph{compensatory phase}'' discussed in \cite{governatori2006formal}, where a secondary clause activates only after the primary obligation fails.
  Thus, our constructions realize these phase shifts ($\mathbf{L} \to \mathbf{R}$) at the automaton level, eliminating the need for the explicit control variables used in prior work.
  \end{remark}



\begin{example}[5-Output Moore Monitors for $C_3$ and $C_2 \wedge C_3$]
\label{ex:moore-c3-literals}
The reparation contract $C_3=\obl[1]{\PAY}\repair\obl[1]{\PAYF}$ combines two obligations:
the primary duty $\obl[1]{\PAY}$ to pay rent,
and the secondary reparation $\obl[1]{\PAYF}$ to pay a late fee if the first is violated.
In addition, $C_2=\perm[1]{\OCC}$ grants the tenant (agent~2) permission to occupy the property.
Below, we show the literal monitors, their reparation composition, and the conjunction $C_2\wedge C_3$,
with verdict outputs $\tightverdicts=\{\mathsf{?},\topt,\bott,\topp,\botp\}$.


We construct $C_2 \wedge C_3$ by the synchronous product of the 5-output monitors for $C_2=\perm[2]{\OCC}$ and $C_3=\obl[1]{\PAY}\repair\obl[1]{\PAYF}$, then keep only
reachable states and minimize construction~\ref{fig:moore-c3-literals}.
The Letter classes used on edges correspond to  shorthand for literal satisfaction or violation conditions:
\[
\begin{aligned}
\PAY^\surd   &:= \{\,A\in\Gamma \mid \{\PAY^{(1)},\PAY^{(2)}\}\subseteq A\,\}, 
&\qquad \PAY^\times   &:= \Gamma\setminus \PAY^\surd,\\[2pt]
\PAYF^\surd  &:= \{\,A\in\Gamma \mid \{\PAYF^{(1)},\PAYF^{(2)}\}\subseteq A\,\},
&\qquad \PAYF^\times  &:= \Gamma\setminus \PAYF^\surd,\\[2pt]
\OCC^{\surd} &:= \{\,A\in\Gamma \mid \OCC^{(2)}\in A \Rightarrow \OCC^{(1)}\in A\,\},
&\qquad \OCC^{\times} &:= \Gamma\setminus \OCC^{\surd}.
\end{aligned}
\]
Outputs follow the conjunction rule $\Lambda$:
$\bott$ if any conjunct tightly rejects, $\botp$ if any conjunct is post-reject,
$\topt$ when one conjunct hits $\topt$ while the other is already at/past acceptance
($\topt$ or $\topp$), $\topp$ if both are post-accept, and $\mathsf{?}$ otherwise.

Reading Subfigure~\ref{fig:c2andc3} with the letter classes defined above: from $s_0$ (pre) there are three behaviors:
(i) If \,$\OCC^\surd \land \PAY^\surd$\ holds in the current letter, both conjuncts succeed (perm is respected, and the primary obligation is met), so the product emits $\topt$
and moves to post-accept $\topp$.
(ii) If \,$\OCC^{\times}$\ holds, the permission conjunct fails tightly, so the product
emits $\bott$ and then $\botp$ forever.
(iii) If \,$\OCC^\surd \land \PAY^{\times}$\ holds, the reparation branch of $C_3$
is activated, and we move to the waiting state $s_1$ (still $\mathsf{?}$). From $s_1$,
\,$\PAYF^\surd$\ discharges the reparation and triggers $\topt$; otherwise
\,$\PAYF^{\times}$\ yields $\bott$. After $\topt$ (accepting frontier), all continuations
are in $\topp$; after $\bott$ (reject frontier) all continuations are in $\botp$.
Thus, the decisive index for the conjunction is the latter of the two successes when both succeed, or the earlier tight failure when any conjunct fails, exactly as the figure shows.



\begin{figure*}[h!]
\centering
\captionsetup[subfigure]{justification=centering}

% ================================================================
% (a)+(b) side-by-side literal monitors
% ================================================================
\begin{subfigure}[t]{0.48\textwidth}
\centering
\scalebox{0.82}{
\begin{tikzpicture}[
  ->,shorten >=1pt, node distance=16mm,
  every state/.style={rectangle,rounded corners,draw,minimum width=10mm,
                      minimum height=6mm,inner sep=4pt,font=\footnotesize,align=center}
]
\node[initial, initial where=above,state,fill=gray!10] (q0) {$s_0$\\$\mathsf{?}$};
\node[state,fill=green!18,right=of q0] (qs) {$s_{\topt}$\\$\topt$};
\node[state,fill=red!18,below=of q0] (qv) {$s_{\bott}$\\$\bott$};
\node[state,fill=green!10,right=of qs] (qps) {$s_{\topp}$\\$\topp$};
\node[state,fill=red!10,below=of qps] (qpv) {$s_{\botp}$\\$\botp$};

\path[->]
  (q0) edge[bend left=10] node[above,pos=0.5] {$\PAY^\surd$} (qs)
  (q0) edge[bend right=10] node[left,pos=0.5] {$ \PAY^\times$} (qv)
  (qs) edge node[above] {$\Gamma$} (qps)
  (qv) edge node[below] {$\Gamma$} (qpv)
  (qps) edge[loop right] node {$\Gamma$} ()
  (qpv) edge[loop right] node {$\Gamma$} ();
\end{tikzpicture}}
\caption{$\tsmc_\ell (\obl[1]{\PAY})$
Satisfaction occurs when both agents perform $\PAY$.}
\label{fig:obl-pay}
\end{subfigure}
\hfill
\begin{subfigure}[t]{0.48\textwidth}
\centering
\scalebox{0.82}{
\begin{tikzpicture}[
  ->, node distance=16mm,
  every state/.style={rectangle,rounded corners,draw,minimum width=10mm,
                      minimum height=6mm,inner sep=4pt,font=\footnotesize,align=center}
]
\node[initial,initial where=above,state,fill=gray!10] (q0) {$s_0$\\$\mathsf{?}$};
\node[state,fill=green!18,right=of q0] (qs) {$s_{\topt}$\\$\topt$};
\node[state,fill=red!18,below=of q0] (qv) {$s_{\bott}$\\$\bott$};
\node[state,fill=green!10,right=of qs] (qps) {$s_{\topp}$\\$\topp$};
\node[state,fill=red!10,below=of qps] (qpv) {$s_{\botp}$\\$\botp$};

\path[->]
  (q0) edge[bend left=10] node[above,pos=0.5] {$\PAYF^\surd$} (qs)
  (q0) edge[bend left=10] node[left,pos=0.5] {$\PAYF^\times$} (qv)
  (qs) edge node[above] {$\Gamma$} (qps)
  (qv) edge node[below] {$\Gamma$} (qpv)
  (qps) edge[loop right] node {$\Gamma$} ()
  (qpv) edge[loop right] node {$\Gamma$} ();
\end{tikzpicture}}
\caption{$\tsmc_\ell (\obl[1]{\PAYF})$
Satisfaction occurs when both agents perform $\PAYF$.}
\label{fig:obl-payf}
\end{subfigure}

\vspace{2mm}

% ================================================================
% (c) Reparation composition
% ================================================================
\begin{subfigure}[t]{0.9\textwidth}
\centering
\scalebox{0.9}{
\begin{tikzpicture}[
  ->,shorten >=1pt, node distance=25mm,
  every state/.style={
    rectangle,rounded corners,draw,
    minimum width=11mm,minimum height=7mm,
    inner sep=5pt,font=\footnotesize,align=center}
]
\node[initial,state,fill=gray!10] (q0) {$s_0$\\$\mathsf{?}$};
\node[state,fill=gray!10,below=of q0] (qwait) {$s_{\mathsf{?2}}$\\[0.2pt]$\mathsf{?}$};
\node[state,fill=green!18,right=of q0] (qs) {$s_{\topt}$\\[0.2pt]$\topt$};
\node[state,fill=red!18,right=of qwait] (qv) {$s_{\bott}$\\[0.2pt]$\bott$};
\node[state,fill=green!10,right=of qs] (qps) {$s_{\topp}$\\[0.2pt]$\topp$};
\node[state,fill=red!10,below=of qps] (qpv) {$s_{\botp}$\\[0.2pt]$\botp$};

\path[->]
  (q0) edge[bend left=10]  node[above,pos=0.45] {\footnotesize$\PAY^{\surd}$} (qs)
  (q0) edge[bend right=10] node[left,pos=0.45]  {\footnotesize$\PAY^{\times}$} (qwait)
  (qwait) edge[bend left=8] node[right,pos=0.45] {\footnotesize$\PAYF^{\surd}$} (qs)
  (qwait) edge[bend right=10] node[below,pos=0.45] {\footnotesize$\PAY^{\times}$} (qv)
  (qs) edge node[above] {\footnotesize$\Gamma$} (qps)
  (qv) edge node[below] {\footnotesize$\Gamma$} (qpv)
  (qps) edge[loop right] node {\footnotesize$\Gamma$} ()
  (qpv) edge[loop right] node {\footnotesize$\Gamma$} ();
\end{tikzpicture}}
\caption{Reparation composition $\tsmc_\repair (\obl[1]{\PAY},\obl[1]{\PAYF})$.
The monitor activates $\PAYF$ after tight failure of $\PAY$.}
\label{fig:obl-pay-repair}
\end{subfigure}

\vspace{2mm}

% ================================================================
% (d) Conjunctive composition (C2 and C3)
% ================================================================
\begin{subfigure}[t]{0.9\textwidth}
\centering
\scalebox{0.9}{
\begin{tikzpicture}[
  ->, node distance=20mm and 18mm,
  every state/.style={
    rectangle,rounded corners,draw,
    minimum width=12mm,minimum height=7mm,
    inner sep=5pt,font=\footnotesize,align=center
  }
]
% States
\node[initial,state,fill=gray!10] (q0) {$s_0$\\[2pt]$\mathsf{?}$};
\node[state, fill=gray!10, below right=16mm and 17mm of q0] (q1) {$s_1$\\[2pt]$\mathsf{?}$};
\node[state,fill=green!18,right=38mm of q0] (qs) {$s_{\topt}$\\[2pt]$\topt$};
\node[state,fill=red!18,below=38mm of q0] (qv) {$s_{\bott}$\\[2pt]$\bott$};
\node[state,fill=green!10,right=38mm of qs] (qps) {$s_{\topp}$\\[2pt]$\topp$};
\node[state,fill=red!10,below=38mm of qps] (qpv) {$s_{\botp}$\\[2pt]$\botp$};

% Transitions
\path[->]
  (q0) edge[bend left=10] node[above,pos=0.5] {\footnotesize$\OCC^\surd \land \PAY^{\surd}$} (qs)
  (q0) edge[bend right=14] node[left,pos=0.5] {\footnotesize$\OCC^{\times}$} (qv)
  (q0) edge[bend left=12] node[pos=0.45,sloped,above] {\footnotesize$\OCC^\surd \land \PAY^{\times}$} (q1)
  (q1) edge[bend left=8] node[right] {\footnotesize$\PAYF^\surd$} (qs)
  (q1) edge[bend left=10] node[sloped,pos=0.5,above] {\footnotesize$\PAYF^{\times}$} (qv)
  (qs) edge node[above] {\footnotesize$\Gamma$} (qps)
  (qv) edge node[below] {\footnotesize$\Gamma$} (qpv)
  (qps) edge[loop right] node {\footnotesize$\Gamma$} ()
  (qpv) edge[loop right] node {\footnotesize$\Gamma$} ();
\end{tikzpicture}}
\caption{Conjunctive composition $\tsmc_{\wedge}(C_2,C_3)$, where $C_2=\perm[1]{\OCC}$.}
\label{fig:c2andc3}
\end{subfigure}


\caption{Literal and composite 5-output Moore monitors for
$C_3=\obl[1]{\PAY}\repair\obl[1]{\PAYF}$.
(a) and (b)~monitors for Literal composing $C_3$  side by side;
(c)~Reparation composition $C_3$;
(d)~Conjunctive composition $C_2 \wedge C_3$,
where $C_2=\perm[1]{\OCC}$ represents the tenant's power to occupy the property. $\OCC^{\surd}:= \{A \mid \{\OCC^{1}\} \not\subset A \sor  \{\OCC^{1},\OCC^{2}\} \subseteq A\}$. 
$\PAY^\surd:= \{A \mid \PAY^{(1)},\PAY^{(2)}\} \subseteq A $. The case $\PAYF^\surd$ is similarly defined as for $\PAY^\surd$.}
\label{fig:moore-c3-literals}
\end{figure*}
\end{example}
\newpage
\subsubsection*{Construction for Binary Regular Expression-Contracts}
Triggered contracts activate their body $C$ once the triggering pattern $re$
reaches its first tight match. Before that point, the monitor behaves exactly as the regular-expression monitor for $re$ and emits only $\mathsf{?}$. Once the trigger fires, the monitor switches permanently to the contract monitor $\tsmc(C)$. If the pattern becomes impossible before it fires, the contract is
vacuously satisfied. The construction follows the same blueprint as sequence and reparation, but applied to the decisive states of the regular expression monitor.

\begin{definition}[Triggered Monitor Construction]
    \label{def:moore-trigger-seq}
    
    \[\text{Let }
    \tsmc(\re)=(Q_r, r_0, \Gamma, \tightverdicts, \delta_r, \lambda_5^r)
    \quad\text{and}\quad
    \tsmc(C)=(Q_c, c_0, \Gamma, \tightverdicts, \delta_c, \lambda_5^c)
    \]
    be five-valued tight satisfaction monitors for the regular expression $\re$ and the contract $C$.
    The triggered monitor for $\trig[re]{C}$ is the machine
    \[
    \tsmc_{\trigg}(\tsmc_{re}(\re),\tsmc(C))
      := (Q_{\trigg}, q_0^{\trigg}, \Gamma, \tightverdicts, \delta_{\trigg}, \lambda_5^{\trigg}).
    \]
    
    \begin{itemize}
      \item The state space includes the open states of the trigger, the states of the contract, and two fresh sink states for vacuous success:
      \[
         Q_{\trigg} := Q_r^{\text{open}} \cup Q_c \cup \{q_{\mathit{vac}}^{\topt}, q_{\mathit{vac}}^{\topp}\},
      \]
      where the reduced regular expressions states are:
      \[
         Q_r^{\text{open}}
         := \{\,q \in Q_r \mid \lambda_5^r(q)\in\{\mathsf{?},\topt,\topp\}\,\}.
      \]
      The initial state is
      \[
        q_0^{\trigg} := r_0.
      \]
    
      \item The transition function $\delta_{\trigg}$ is defined in three parts:
    
      \begin{enumerate}
          \item 
      \textbf{Guard-active region ($q\in Q_r^{\text{open}}$).}
      Let $q'=\delta_r(q,A)$.
      \[
      \delta_{\trigg}(q,A)=
      \begin{cases}
        c_0
          & \text{if } q'\in Q_r^{\topt}
            \quad\text{(first tight match: activate } C\text{)},\\[4pt]
        q_{\mathit{vac}}^{\topt}
          & \text{if } q'\in Q_r^{\bott}
            \quad\text{(trigger impossible: vacuous success)},\\[4pt]
        q'
          & \text{if } q'\in Q_r^{?}
            \quad\text{(guard still open)}.
      \end{cases}
      \]
      
      \item  \textbf{Contract-active region ($y\in Q_c$).} 
      \[
        \delta_{\trigg}(y,A) := \delta_c(y,A).
      \]
      
      \item \textbf{Vacuous success sinks.}
      \[
        \delta_{\trigg}(q_{\mathit{vac}}^{\topt}, A) := q_{\mathit{vac}}^{\topp}, \qquad
        \delta_{\trigg}(q_{\mathit{vac}}^{\topp}, A) := q_{\mathit{vac}}^{\topp}.
      \]
      \end{enumerate}
    
      \item The output function is
      \[
      \lambda_5^{\trigg}(q)=
      \begin{cases}
        \mathsf{?}
          & \text{if } q\in Q_r^{\text{open}},\\[2pt]
        \lambda_5^{c}(q)
          & \text{if } q\in Q_c,\\[2pt]
        \topt
          & \text{if } q = q_{\mathit{vac}}^{\topt},\\[2pt]
        \topp
          & \text{if } q = q_{\mathit{vac}}^{\topp}.
      \end{cases}
      \]
    \end{itemize}
    \end{definition}
    
    
    \paragraph{Intuition.}
    The trigger monitor is obtained with the same redirection recipe used for sequence, but applied to the trigger.
    Keep only states that are still open in the trigger (outputs in \(\{\,\mathsf{?},\topt,\topp\,\}\)). Remove its decisive states.
    Redirect every transition that would be the first tight match of the triggering regular expression (the step that enters \(\topt\)) to the initial state of the contract monitor \(C\).
    From that point on, the global output is exactly the output of \(C\) on the suffix.
    Redirect every transition that would make the trigger impossible (enter \(\bott\) or \(\botp\)) to the fresh vacuous-success sink \(q_{\mathit{vac}}^{\topt}\), which emits \(\topt\) and then permanently \(\topp\).
    While the guard remains open, only the guard component advances and the product emits \(\mathsf{?}\), so no premature verdict appears.
    This realizes the prefix clauses: success either because the trigger never becomes true (vacuous satisfaction), or because it fires at the earliest index (matching the tight prefix) and the suffix satisfies \(C\);
    violation only if the guard fires and the suffix violates \(C\).
    
    \begin{lemma}[Correctness of the triggered monitor construction]
    \label{lem:trigg-correct}
    \[\text{Let } \tsmc_{\trigg}(\tsmc_{re}(\re),\tsmc(C))
          =(Q_{\trigg},\Gamma,\delta_{\trigg},q_0^{\trigg},\lambda_5^{\trigg})\]
    be the monitor constructed in Definition~\ref{def:moore-trigger-seq} for
    $\trig[re]{C}$.
    Then, for every trace $\pi$,
    \[
    \lambda_5^{\trigg}\bigl(\delta_{\trigg}(q_0^{\trigg},\pi)\bigr)
    \;=\;
    \semfive{\pi \vDash \trig[re]{C}}.
    \]
    \end{lemma}
    
    \noindent
    \textbf{Proof sketch.}
    The proof follows the same pattern as for sequence and reparation, by induction on the length of $\pi$.
    
    \smallskip
    \emph{Base case.}
    For $\pi=\emptytrace$ the monitor is in $q_0^{\trigg}=r_0$, the initial state
    of the guard.
    The value
    \[
    \lambda_5^{\trigg}\bigl(\delta_{\trigg}(q_0^{\trigg},\emptytrace)\bigr)
      = \lambda_5^r(r_0) = \mathsf{?}
    \]
    coincides with the tight semantics of $\trig[re]{C}$ on the empty trace: no trigger has fired, and no violation has occurred.
    
    \smallskip
    \emph{Inductive step.}
    Assume the invariant holds for all prefixes of length $n$.
    Consider a prefix
    of length $n{+}1$ and its last letter $A$.
    There are three regions:
    
    \begin{itemize}
      \item \emph{Guard-active region ($x\in Q_r^{\text{open}}$).}
      By construction, $\delta_{\trigg}(x,A)$ is:
      \begin{itemize}
        \item $c_0$, if $\delta_r(x,A)\in Q_r^{\topt}$.
        This is exactly the case
        where $re$ reaches tight success for the first time.
        The next state is the
        initial state of $\tsmc(C)$, so subsequent behavior matches the semantics
        of $C$ on the suffix.
        This realizes the clause “trigger fires at the
        earliest index and the suffix must satisfy $C$”.
        
        \item $q_{\mathit{vac}}^{\topt}$, if $\delta_r(x,A)\in Q_r^{\bott}$, that is, the pattern becomes impossible.
        This state outputs $\topt$ and transitions to the permanent success state $q_{\mathit{vac}}^{\topp}$.
        This matches the semantics where the trigger never fires and $\trig[re]{C}$
        holds vacuously (tight success followed by post-success).
    
        \item $\delta_r(x,A)$ if $\delta_r(x,A)\in Q_r^{?}$, in which case the
        guard remains open, and the global verdict stays $\mathsf{?}$.
        This matches the semantic clause that no decisive information is available as long as neither a match nor an impossibility has been detected.
      \end{itemize}
      The inductive hypothesis on the guard monitor ensures that the moment of redirection coincides with the earliest decisive prefix of $re$.
    
      \item \emph{Contract-active region ($y\in Q_c$).}
      Once the monitor has been redirected into $c_0$, all transitions are given by $\delta_c$, and outputs by $\lambda_5^c$.
      The induction hypothesis for
      $\tsmc(C)$ gives
      \[
      \lambda_5^{\trigg}\bigl(\delta_{\trigg}(q_0^{\trigg},\pi)\bigr)
       = \lambda_5^{c}\bigl(\delta_c(c_0,\pi')\bigr)
       = \semfive{\pi' \vDash C},
      \]
      where $\pi'$ is the suffix after the trigger point.
      This matches the
      semantics of $\trigg[re]{C}$ on all traces where the trigger has fired.
      
      \item \emph{Vacuous success region.}
      If the monitor enters $q_{\mathit{vac}}^{\topt}$ or $q_{\mathit{vac}}^{\topp}$, it correctly emits the success verdicts corresponding to a triggered contract whose trigger can no longer be satisfied.
    \end{itemize}
    
    \smallskip
    \emph{Violation and satisfaction cases.}
    If the trigger fires at some earliest index $k$ and the suffix $\pi^{k+1}$
    violates $C$ tightly or post, the monitor is in the contract-active region and
    outputs the corresponding $\bott$ or $\botp$, which is exactly
    $\semfive{\pi \vDash \trig[re]{C}}$ in this case.
    If the trigger never fires and the guard becomes impossible, the monitor
    outputs $\topt$ and then $\topp$, which matches vacuous satisfaction.
    In all other cases the output remains $\mathsf{?}$, as the semantics of
    $\trigg[re]{C}$ leaves the status undecided.
    
    \smallskip
    \emph{Conclusion.}
    At each prefix, the monitor either simulates the guard with the correct decisive redirection points, simulates $C$ on the correct suffix, or enters the correct vacuous success state.
    Hence
    for every trace $\pi$ the monitor verdict
    \(
    \lambda_5^{\trigg}(\delta_{\trigg}(q_0^{\trigg},\pi))
    \)
    coincides with the five-valued tight semantics of $\trig[re]{C}$.
    \qed





\begin{definition}[Guarded Monitor Construction]
\label{def:moore-guard}
\[\text{Let }
\tsmc(\re)=(Q_r, r_0, \Gamma, \tightverdicts, \delta_r, \lambda_5^r)
\quad\text{and}\quad
\tsmc(C)=(Q_c, c_0, \Gamma, \tightverdicts, \delta_c, \lambda_5^c).
\]
The guarded contract monitor for $\guard[re]{C}$ is the synchronous product
\[
\tsmc_{\guardd}(\tsmc_{re}(\re),\tsmc(C))
  :=(Q_r\times Q_c,\ (r_0,c_0),\ \Gamma,\ \tightverdicts,\ \delta_{\guardd},\ \lambda_5^{\guardd}),
\]
with
\[
\delta_{\guardd}((x,y),A)
  :=(\delta_r(x,A),\delta_c(y,A)).
\]
The output verdict of the guarded monitor is determined by jointly inspecting the current verdicts of the guard and the contract components, according to the following case distinction.
\[
\lambda_5^{\guardd}(x,y)=
\begin{cases}
\bott
  & \text{if }\lambda_5^r(x)\in\{\mathsf{?},\topt,\topp\}
    \ \text{and }\lambda_5^c(y)=\bott,\\[2pt]
\botp
  & \text{if }\lambda_5^r(x)\in\{\mathsf{?},\topt,\topp\}
    \ \text{and }\lambda_5^c(y)=\botp,\\[2pt]
\topt
  & \text{if }\lambda_5^r(x)=\bott
    \ \text{and }\lambda_5^c(y)\in\{\mathsf{?},\topt,\topp\},\\[2pt]
\topp
  & \text{if }\lambda_5^r(x)=\botp
    \ \text{and }\lambda_5^c(y)\in\{\mathsf{?},\topt,\topp\},\\[2pt]
\topt
  & \text{if }\lambda_5^r(x)\in\{\topt,\topp\}
    \ \text{and }\lambda_5^c(y)=\topt,\\[2pt]
\topp
  & \text{if }\lambda_5^r(x)\in\{\topt,\topp\}
    \ \text{and }\lambda_5^c(y)=\topp,\\[2pt]
\mathsf{?}
  & \text{otherwise.}
\end{cases}
\]
\end{definition}



\paragraph{Intuition.}
The guard reads both monitors in lockstep and enforces:

\begin{itemize}
  \item \emph{Open guard.} While the guard is open
  \(\bigl(\lambda_r\in\{\mathsf{?},\topt,\topp\}\bigr)\), i.e. exactly when \(\pi\ \clossat\ re\),
  any tight or post failure of \(C\) becomes the global failure:
  \begin{align*}
    \pi\ \violt\ \guard[re]{C} &\iff (\pi\ \clossat\ re)\ \land\ (\pi\ \violt\ C),\\
    \pi\ \postviol\ \guard[re]{C} &\iff (\pi\ \clossat\ re)\ \land\ (\pi\ \postviol\ C).
  \end{align*}

  \item \emph{Guard impossible.} If the guard becomes impossible
  \(\bigl(\lambda_r\in\{\bott,\botp\}\bigr)\), we accept provided \(C\) has not failed:
  \begin{align*}
    \pi\ \satt\ \guard[re]{C} &\iff (\pi\ \violt\ re)\ \land\ (\pi\ \clossat\ C),\\
    \pi\ \postsat\ \guard[re]{C} &\iff (\pi\ \violt\ re)\ \land\ (\pi\ \postsat\ C).
  \end{align*}

  \item \emph{Guard fired/closed.} When the guard has fired/closed
  \(\bigl(\lambda_r\in\{\topt,\topp\}\bigr)\), we require \(C\) to (tight/post) succeed:
  \begin{align*}
    \pi\ \satt\ \guard[re]{C} &\iff (\pi\ \clossat\ re)\ \land\ (\pi\ \satt\ C),\\
    \pi\ \postsat\ \guard[re]{C} &\iff (\pi\ \clossat\ re)\ \land\ (\pi\ \postsat\ C).
  \end{align*}
\end{itemize}
Note that $\pi\ \clossat\ re$ holds exactly when $\lambda_5^r\in{\mathsf{?},\topt,\topp}$.
The resulting case table is symmetric and total, and it collapses to the expected two-valued clauses once tight/post outcomes are merged into satisfied vs.\ violated.

\begin{lemma}[Correctness of the guarded monitor construction]
\label{lem:guard-correct}
\[\text{Let }
\tsmc_{\guardd}(\tsmc_{re}(\re),\tsmc(C))
  =(Q_{\guardd},q_0^{\guardd},\Gamma,\tightverdicts,\delta_{\guardd},\lambda_5^{\guardd})
\]
be the monitor constructed in Definition~\ref{def:moore-guard} for
$\guard[re]{C}$.  
Then, for every finite trace $\pi$,
\[
\lambda_5^{\guardd}\bigl(\delta_{\guardd}(q_0^{\guardd},\pi)\bigr)
  \;=\;
\semfive{\pi \vDash \guard[re]{C}}.
\]
\end{lemma}
\noindent\textbf{Proof sketch.}
The argument proceeds by induction on the length of $\pi$.  
The guarded monitor is a synchronous product of $\tsmc(re)$ and $\tsmc(C)$,
with the output governed by the case distinction in
Definition~\ref{def:moore-guard}.  
Each region of the case table matches exactly one of the semantic clauses for 
$\guard[re]{C}$.

\smallskip
\emph{Base case.}
For $\pi=\emptytrace$ we have
\[
\lambda_5^{\guardd}\bigl(\delta_{\guardd}(q_0^{\guardd},\emptytrace)\bigr)
  = \lambda_5^{\guardd}(r_0,c_0),
\]
which yields $\mathsf{?}$ in agreement with
$\semfive{\emptytrace \vDash \guard[re]{C}}$.

\smallskip
\emph{Inductive step.}
Assume correctness for all prefixes of length $n$.
Consider $\pi[n{+}1]$ with last letter $A$ and let
\[
(x',y') := \delta_{\guardd}((x,y),A)
           = (\delta_r(x,A),\delta_c(y,A)).
\]

There are three semantic regions, corresponding to the three guard statuses.

\begin{itemize}
  \item \textbf{Guard open}
  \(\lambda_5^r(x)\in\{\mathsf{?},\topt,\topp\}\).  
  This means $\pi$ still possibly satisfies the guard.  
  The guarded semantics requires that any tight or post failure of $C$ becomes the global failure.  
  The monitor table assigns $\bott$ or $\botp$ precisely in these cases, and
  $\mathsf{?}$ otherwise, matching
  \[
  \semfive{\pi \vDash \guard[re]{C}}
    = \semfive{\pi \vDash C}
    \quad\text{as long as $re$ is still open.}
  \]

  \item \textbf{Guard impossible}
  \(\lambda_5^r(x)\in\{\bott,\botp\}\).  
  This corresponds exactly to $\pi\ \violt\ re$.  
  The guarded semantics declares vacuous acceptance provided that $C$ has not already failed.  
  The output table assigns $\topt$ or $\topp$ if $C$ has not failed, and
  propagates $\bott$ or $\botp$ if it has.  
  This matches the semantic requirements for vacuous satisfaction.

  \item \textbf{Guard closed (triggered or concluded)}
  \(\lambda_5^r(x)\in\{\topt,\topp\}\).  
  In this region, the guard has fired or completed successfully, and the
  semantics require $C$ to satisfy or to fail.  
  The output table combines the post-accept and tight-accept verdicts of $C$
  with those of $re$ exactly as demanded by the five-valued semantics:
  \[
     \semfive{\pi \vDash \guard[re]{C}}
       = \semfive{\pi \vDash C}
       \quad\text{once $re$ has closed.}
  \]
\end{itemize}

\smallskip
\emph{Conclusion.}
At each prefix of $\pi$, the guarded monitor outputs exactly the verdict
prescribed by the five-valued semantics of $\guard[re]{C}$.  
Thus
\[
\lambda_5^{\guardd}\bigl(\delta_{\guardd}(q_0^{\guardd},\pi)\bigr)
  = \semfive{\pi \vDash \guard[re]{C}}
\]
for all traces $\pi$.
\qed







\subsubsection*{Construction for repetition contracts}

Repetition contracts describe behaviors that must be satisfied several times in
sequence. The monitor construction follows the intuition that each repetition runs an independent copy of the monitor for $C$, with the next copy becoming active exactly when the previous one reaches tight success. For finite
repetition $C^n$, this results in $n$ chained monitors. For unbounded
repetition $\repit{C}$, we obtain an infinite cycle without ever reporting
tight success.

\medskip
The following constructions make these ideas explicit.
\begin{definition}[Tight monitor construction for finite repetition $\tsmc_\nrep(n,C)$]
\label{def:moore-repeat-finite}
\[\text{Let } \tsmc(C)=(Q,q_0,\Gamma,\tightverdicts,\delta,\lambda_5)\]
be the five-valued tight satisfaction monitor for $C$.  
For $n\in\mathbb{N}^*$, the tight monitor for the finite repetition contract
$\tsmc_\nrep(n,C)$ is defined as
\[
\tsmc_{\nrep}(n,C)
  := (Q_{\nrep},q_0^{\nrep},\Gamma,\tightverdicts,\delta_{\nrep},\lambda_5^{\nrep}).
\]

\medskip
\noindent\emph{Disjoint copies.}
For each $i\in\{1,\dots,n\}$, create a disjoint copy of the base monitor:
\[
Q^{(i)}=\{q^{(i)}\mid q\in Q\},\qquad
\delta^{(i)}(q^{(i)},A)=(\delta(q,A))^{(i)},\qquad
\lambda_5^{(i)}(q^{(i)})=\lambda_5(q).
\]
Write $q_0^{(i)}$ for the copy of $q_0$.

\medskip
\noindent\emph{State space and start state.}
\[
Q_{\nrep}
  := \Bigl(\bigcup_{i=1}^n Q^{(i)}\Bigr)\cup\{q_{\topp},q_{\botp}\},
\qquad
q_0^{\nrep} := q_0^{(1)},
\]
where $q_{\topp}$ and $q_{\botp}$ are fresh post-success and post-failure sinks.

\medskip
\noindent\emph{Transition function.}  
For $1\le i\le n-1$:
\[
\delta_{\nrep}(q^{(i)},A)=
\begin{cases}
q_0^{(i+1)}
   & \text{if }\lambda_5^{(i)}(\delta^{(i)}(q^{(i)},A))=\topt,\\[2pt]
\delta^{(i)}(q^{(i)},A)
   & \text{if }\lambda_5^{(i)}(\delta^{(i)}(q^{(i)},A))
      \notin\{\topt,\topp\}.
\end{cases}
\]
For the last copy $i=n$:
\[
\delta_{\nrep}(q^{(n)},A)=
\begin{cases}
q_{\topp}
   & \text{if }\lambda_5^{(n)}(\delta^{(n)}(q^{(n)},A))=\topt,\\[2pt]
\delta^{(n)}(q^{(n)},A)
   & \text{if }\lambda_5^{(n)}(\delta^{(n)}(q^{(n)},A))
      \notin\{\topt,\topp\}.
\end{cases}
\]
Sink states absorb:
\[
\delta_{\nrep}(q_{\topp},A)=q_{\topp},
\qquad
\delta_{\nrep}(q_{\botp},A)=q_{\botp}.
\]

\medskip
\noindent\emph{Output function.}
For $q^{(i)}\in Q^{(i)}$:
\[
\lambda_5^{\nrep}(q^{(i)})=
\begin{cases}
\bott & \text{if }\lambda_5^{(i)}(q^{(i)})=\bott,\\
\mathsf{?} & \text{if }\lambda_5^{(i)}(q^{(i)})
            \in\{\mathsf{?},\topt,\topp\}.
\end{cases}
\]
For sink states:
\[
\lambda_5^{\nrep}(q_{\topp})=\topp,
\qquad
\lambda_5^{\nrep}(q_{\botp})=\botp.
\]

\medskip
\noindent
This completes the construction of the five-valued monitor for $C^n$.
\end{definition}

\begin{lemma}[Correctness of the finite repetition monitor]
\label{lem:repeat-finite-correct}
For every finite trace $\pi$, contract $C$, and
$\tsmc_{\nrep}(n,C)
  := (Q_{\nrep},q_0^{\nrep},\Gamma,\tightverdicts,\delta_{\nrep},\lambda_5^{\nrep})$, the following holds
\[
\lambda_5^{\nrep}\bigl(\delta_{\nrep}(q_0^{\nrep},\pi)\bigr)
   = \semfive{\pi \vDash C^n}.
\]
\end{lemma}




In the next sections, we use the definitions of the different languages in the semantics of \cdl, along with automata constructions and transformations, to enable automatic detection of violations and the attribution of blame for synchronous interactions over contracts in \cDL.


\begin{definition}[Tight monitor construction for unbounded repetition $\repit{C}$]
\label{def:moore-repeat-unbounded}
Let
\[
\tsmc(C)=(Q,q_0,\Gamma,\tightverdicts,\delta,\lambda_5)
\]
be the tight satisfaction monitor for $C$.
The monitor for the unbounded repetition $\repit{C}$ is defined as
\[
\tsmc_{\text{Rep}}
  := (Q_\omega,q_0,\Gamma,\tightverdicts,\delta_\omega,\lambda_5^{\omega})
\]
where the construction removes all success states of $C$ and redirects
tight success back to the initial state.

\medskip
\noindent\emph{State space.}
\[
Q_\omega := Q \setminus Q_{\top},
\qquad
Q_{\top} := \{\, q\in Q \mid \lambda_5(q)\in\{\topt,\topp\}\,\}.
\]

\medskip
\noindent\emph{Transition function.}
For every $q\in Q_\omega$ and $A\in\Gamma$:
\[
\delta_\omega(q,A)=
\begin{cases}
q_0
  & \text{if }\lambda_5(\delta(q,A))=\topt 
    \quad(\text{restart next iteration after tight success}),\\[2pt]
\delta(q,A)
  & \text{otherwise}.
\end{cases}
\]

\medskip
\noindent\emph{Output function.}
The monitor never declares satisfaction:
\[
\lambda_5^{\omega}(q)=
\begin{cases}
\bott & \text{if }\lambda_5(q)=\bott,\\[2pt]
\mathsf{?} & \text{if }\lambda_5(q)\in\{\mathsf{?},\topt,\topp\}.
\end{cases}
\]

This yields the tight five-valued monitor for $\repit{C}$.
\end{definition}

\begin{lemma}[Correctness of the unbounded repetition monitor]
\label{lem:repeat-unbounded-correct}
For every finite trace $\pi$, and
$\tsmc_{\repit{}}(C)
  := (Q_\omega,q_0,\Gamma,\tightverdicts,\delta_\omega,\lambda_5^{\omega})$
, the following holds
\[
\lambda_5^{\omega}\bigl(\delta_\omega(q_0,\pi)\bigr)
  = \semfive{\pi \vDash \repit{C}}.
\]
\end{lemma}
The monitor never emits $\topt$ nor $\topp$, because tight satisfaction of
$\repit{C}$ is false.  
It emits $\bott$ (and then permanently $\botp$) exactly when some iteration of
$C$ violates tightly, i.e.
\[
\exists n\in\mathbb{N}:\ \pi \violt C^n.
\]
Otherwise, it remains in the undecided verdict $\mathsf{?}$.



\begin{example}[Open ended contract monitor construction]
\label{ex:open-ended}

We illustrate the constructions for the guarded open-ended contract \(\guard[re]{\repit{C_3}}\) where
\(re = \Gamma^{+};\{\notifterm^{(1)}\}\) and
\(C_3=\obl[1]{\PAY}\repair\obl[1]{\PAYF}\).
The construction proceeds in three steps:
the unbounded-repetition monitor for \(C_3\),
the regular-expression monitor for \(re\),
and finally the guarded product. The resulting automata
are shown in \ref{fig:one-per-line}.

\medskip
\noindent\textbf{(a) \(\tsmc(\repit{C_3})\)
(Figure~\ref{fig:rep-c3_vertical}).}
The monitor contains the states \(q_0/\mathsf{?}\), \(q_w/\mathsf{?}\), \(q_{\bott}/\bott\), and \(q_{\botp}/\botp\). The meaning of the transitions is as follows.

A joint payment \(\PAY^\surd\) keeps the monitor at \(q_0\).
A missed payment \(\PAY^\times\) moves to the waiting state \(q_w\).
From \(q_w\), a successful late fee \(\PAYF^\surd\) restarts the cycle
by returning to \(q_0\). A failed repair \(\PAYF^\times\) produces a tight
violation and moves to \(q_{\bott}\), which then steps on any letter to the
permanent sink \(q_{\botp}\). The sink loops on all letters.

This matches the construction of Definition~\ref{def:moore-repeat-unbounded}:
tight success restarts a new cycle, and tight failure leads to a permanent
violation.

\medskip
\noindent\textbf{(b) \(\tsmc_{re}(re)\) for
\(re = \Gamma^{+};\{\notifterm^{(1)}\}\)
(Figure~\ref{fig:guard-}).}
The monitor has states \(s_0/\mathsf{?}\), \(s_1/\mathsf{?}\), \(s^t/\topt\), and \(s^{+}/\topp\).

The regular-expression part reads arbitrary letters:
\(s_0 \xrightarrow{*} s_1\).
While no termination notice is received, the machine remains in \(s_1\)
via \(\overline{T} = \Gamma \setminus T\).
A letter in \(T=\{A\mid \notifterm^{(1)}\in A\}\) produces a tight match
and moves to \(s^t\). Any continuation moves to the post-acceptance state
\(s^{+}\), which loops on all letters.

\medskip
\noindent\textbf{(c) Guarded product
\(\tsmc_\guardd(\tsmc_{re}(re),\tsmc(\repit{C_3})\)
(Figure~\ref{fig:guard-repC3}).}
The product is organised by verdict class:
violating states on the left, undecided states in the centre, and
accepting states on the right.

\smallskip
\emph{Undecided region (centre).}
The reachable combinations while the guard is open are
\(s_0\times q_0\),
\(s_1\times q_w\),
and \(s_1\times q_0\).
Their transitions follow the product rule \((x,y)\xrightarrow{A}(\delta_r(x,A),\delta_c(y,A))\).
Examples include:
\[
\begin{aligned}
&s_0\times q_0 \xrightarrow{\PAY^\times} s_1\times q_w,\\
&s_1\times q_w \xrightarrow{\PAYF^\times} s_1\times q_{\bott},\\
&s_1\times q_w \xrightarrow{\PAYF^\surd \wedge \overline{T}} s_1\times q_0,\\
&s_1\times q_0 \xrightarrow{\PAY^\times \wedge \overline{T}} s_1\times q_w,\\
&s_0\times q_0 \xrightarrow{\PAY^\surd} s_1\times q_0.
\end{aligned}
\]

\smallskip
\emph{Violation region (left).}
Once the contract component reaches \(q_{\bott}\),
the guard is still open so the product outputs \(\bott\) and moves on any
letter to \(S\times q_{\botp}\), which loops on every letter.

\smallskip
\emph{Acceptance region (right).}
When the guard fires on a letter in \(T\), the product moves to
\(s^t\times(q_0,q_w)\) with verdict \(\topt\) provided the repetition contract has
not violated. From there, any continuation leads to the post-acceptance
state \(s^{+}\times Q\) that loops on all letters and emits \(\topp\).

\medskip
\noindent
This behavior follows directly from Definition~\ref{def:moore-guard}:
while \(\lambda_r\in\{\mathsf{?},\topt,\topp\}\) the guard is open, so any tight or
post violation of \(\repit{C_3}\) becomes a violation of the guarded contract.
Once the guard becomes true on \(T\), the contract must satisfy \(\repit{C_3}\)
from that point on, for the global verdict to be tight or post acceptance.


\end{example}


\begin{figure}[h!]
\centering
\tikzset{
  ->,shorten >=1pt, semithick,
  node distance=17mm,
  every state/.style={
    rectangle,rounded corners,draw,
    minimum width=12mm,minimum height=7mm,
    inner sep=5pt,font=\footnotesize,align=center}
}

% M(rep(C3)) — full width
\begin{subfigure}[t]{0.94\textwidth}
\centering
\begin{tikzpicture}
  \node[initial,state,fill=gray!10]          (q0)  {$q_0$\\$\mathsf{?}$};
  \node[state,fill=gray!10,below=3cm of q0]      (qw)  {$q_w$\\$\mathsf{?}$};
  \node[state,fill=red!18,right=3cm of qw]       (qv)  {$q_{\bott}$\\$\bott$};
  \node[state,fill=red!10,above= 3cm of qv]       (qpv) {$q_{\botp}$\\$\botp$};

  \path
    % In-cycle behavior:
    (q0) edge[loop above] node[above,pos=0.5] {\footnotesize $\PAY^\surd$} ()
    (q0) edge[bend right=13] node[left,pos=0.45] {\footnotesize $\PAY^\times$} (qw)
    (qw)  edge[bend right=13] node[right,pos=0.45] {\footnotesize $\PAYF^\surd$} (q0) % restart on tight success
    (qw)  edge[bend right=10] node[below,pos=0.45] {\footnotesize $\PAYF^\times$} (qv)
    % Failure sink:
    (qv)  edge node[left] {$\Gamma$} (qpv)
    (qpv) edge[loop right] node {$\Gamma$} ();
\end{tikzpicture}
\caption{Monitor for $\repit{C_3}$.}
\label{fig:rep-c3_vertical}
\end{subfigure}



\vspace{3mm}
\begin{subfigure}[t]{0.94\textwidth}
\centering
\begin{tikzpicture}
\node[initial,state,fill=gray!10]  (s0)  {$s_0$\\[0.2pt]$\mathsf{?}$};
\node[state,fill=gray!10,right=of s0]  (g)   {$s_1$\\[0.2pt]$\mathsf{?}$};
\node[state,fill=green!18,right=16mm of g] (acc) {$s^t$\\[0.2pt]$\topt$};
\node[state,fill=green!10,right=19mm of acc] (ap) {$s^{+}$\\[0.2pt]w$\topp$};

% --- Transitions ---
\path
  (s0) edge node[above,pos=0.4] {$\Gamma$} (g)
  (g) edge[loop below] node[below] {$\overline{T}$} ()
  (g) edge node[above] {$T$} (acc)
%   (t2) edge node[left,pos=0.5] {$\Gamma$} (acc)
  (acc) edge node[above] {$\Gamma$} (ap)
  (ap) edge[loop above] node {$\Gamma$} ();

\end{tikzpicture}
\caption{Monitor for the regular expression  $\Gamma^+ \cdot  \ \{terminate^{(1)}\}$.}
\label{fig:guard-}
\end{subfigure}
\vspace{3mm}

% -----------------------------------------
% (b) Guarded: guard[re_C5]{rep(C3)} — full width
% -----------------------------------------


\begin{subfigure}[t]{0.94\textwidth}
\centering
\begin{tikzpicture}

% --- Columns: Bad (x=-3) | ? (x=0) | Accept (x=+3) ---

% ? column (Open, undecided)
\node[initial,state,fill=gray!10]  (Oq0)  at (0, 0) {$s_0\times q_0$\\[0.2pt]$\mathsf{?}$};
\node[state,fill=gray!10]          (Oqw)  at (0,-2.6) {$s_1\times q_w$\\[0.2pt]$\mathsf{?}$};

\node[state,fill=gray!10]  (Oqq00)  at (4, 0) {$s_1\times q_0$\\[0.2pt]$\mathsf{?}$};
%\node[state,fill=gray!10]          (Oqw0)  at (3,-2) {$\text{Open}\times q_w$\\$\mathsf{?}$};

% Bad column (left): tight/post violation sinks
\node[state,fill=red!18]           (Oqv)  at (-3.4,-2.6) {$(s_0,s_1)\times q_{\bott}$\\[0.2pt]$\bott$};
\node[state,fill=red!10]           (Oqpv) at (-3.4, 0) {$S \times q_{\botp}$\\[0.2pt]$\botp$};

% Accept column (right): guard impossible (vacuous accept)
\node[state,fill=green!18]         (Iq0t) at (4, -2.6) {$s^t\times (q_0,q_w)$\\[0.2pt]$\topt$};
\node[state,fill=green!10]         (Iqwt) at (6.4,-2.6) {$s^+\times Q$\\[0.2pt]$\topp$};

% --- Open-region transitions (guard open; C repeats) ---
\path
  %(Oq0) edge[loop above] node {\scriptsize $\overline T$ or $\PAY^\surd$} ()
  (Oq0) edge node[left, inner sep=2pt]{\footnotesize$\PAY^\times$} (Oqw)
  (Oqw) edge[bend left=14] node[below, inner sep=2pt]{\footnotesize $\PAYF^\times$} (Oqv)
 % (Oqw) edge[bend left=12] node[left]{\scriptsize $\PAYF^\surd$} (Oq0)
  (Oqv) edge              node[left] {\footnotesize$\Gamma$} (Oqpv)
  (Oqpv) edge[loop left]  node {\footnotesize$\Gamma$} ();

% --- Mode switches: Open -> Imposs (guard hits bott/botp) ---
\path
  (Oq0) edge[bend left=10]  node[above, inner sep=2pt]{\footnotesize$\PAY^\surd$} (Oqq00)
  (Oqw) edge [bend left=7] node[sloped,above, inner sep=2pt,pos=0.4]{\footnotesize $\PAYF^\surd \wedge \overline{T}$} (Oqq00)
  (Oqq00) edge  [bend left=7] node[sloped,below,inner sep=2pt,pos=0.4] {\footnotesize $\PAY^\times \wedge \overline{T}$} (Oqw)
  
  (Oqq00) edge  node[right] {$T$} (Iq0t)
  (Oqw) edge[bend right=12]  node[below, inner sep=2pt]{\footnotesize $\PAYF^\surd \wedge T$} (Iq0t)
  (Iq0t) edge[bend right=12]  node[below, inner sep=2pt]{\footnotesize $\Gamma$} (Iqwt);
  %(Oq0) edge[bend left=14]  node[below,sloped] {\scriptsize guard $\botp$} (Iqwt)
  %(Oqw) edge[bend left=16]  node[below,sloped] {\scriptsize guard $\botp$} (Iqwt);

% --- Imposs region loops ---
\path
 % (Iq0t) edge[loop right] node {$\Gamma$} ()
  (Iqwt) edge[loop right] node {\footnotesize$\Gamma$} ()
  (Oqq00) edge[loop right] node {\footnotesize$\PAY^\surd \wedge \overline{T}$} ();

\end{tikzpicture}
\caption{Tight satisfaction monitor construction for $\guard[\Gamma^+ \cdot \notifterm^{(1)}]{\repit{C_3}}$.}
\label{fig:guard-repC3}
\end{subfigure}
\caption{%
Progressive construction for $\guard[(\Gamma^+ \cdot \notifterm^{(1)})]{\repit{C_3}}$.
Figure (c) is obtained by applying $\tsmc_\guardd$ to Figures (b) and (a).
Here, roughly speaking, $\PAY^\surd := \{\,A\in\Gamma \mid \{\PAY^{(1)},\PAY^{(2)}\}\subseteq A\,\}$ denotes a joint payment event,$\PAY^\times := \Gamma \setminus \PAY^\surd $ denotes failure of payment.
$T := \{\,A\in\Gamma \mid \notifterm^{(1)}\in A\,\}$ denotes the occurrence of a termination notice by agent~1,
and $\overline{T} := \Gamma\setminus T$ denotes its absence.
}
\label{fig:one-per-line}
\end{figure}



The guarded open-ended contract illustrates how the automaton constructions combine.  
The unbounded repetition of \(C_3\) enforces an indefinite sequence of payments with repair, and the regular expression \(re\) recognizes the first occurrence of a termination notice.  
The guarded product synchronizes both behaviors: while the guard is open, all failures of \(\repit{C_3}\) propagate to the global verdict; once the guard closes, the open-ended obligation is discharged, and the monitor collapses to post acceptance provided that no violation has occurred.  
This example shows that the tight constructions scale to nested patterns of sequencing, repetition, guarding, and reparation while still preserving a direct correspondence with the prefix semantics of \cDL.