\chapter{Preliminaries}\label{chapter:preliminaries}
This chapter introduces the mathematical structures that underlie the temporal component of our normative framework, together with the basic models used to represent system executions. We focus on representations that allow timed constraints to be stated, combined, and analyzed symbolically, without committing to a particular execution or monitoring model. In parallel, we introduce trace-based word models that serve as the semantic carriers for observed behavior and interaction. The material in this chapter is intentionally elementary and self contained. Its purpose is not to introduce new logical operators, but to establish precise domains for manipulating time ranges and executions algebraically, which will later be embedded into normative specifications and monitoring constructions.
\section{Interval Set over Integers}
Timed constraints in normative systems are rarely punctual. Obligations and prohibitions typically apply over contiguous or fragmented time ranges, and reasoning directly over individual time points leads to unnecessary blow up. For this reason, we adopt sets of intervals as our basic representation of temporal constraints. Interval sets provide a compact and canonical way to capture possibly discontinuous time ranges, while still supporting precise arithmetic operations such as intersection, union, and difference. Before introducing interval sets and their algebra, we first fix the underlying numerical domain on which these intervals are defined.
\subsection{Natural Numbers and Extended Natural Numbers}

Let  $\mathbb{N}$ denote the set of all positive integers including zero, i.e.\ \( \mathbb{N}:= \{ 0, 1, 2, 3, \dots \} \) and let \emph{less} ($<$) and  \emph{less or equal} ($\leq$) denote the usual order relations on $\mathbb{N}$.
%
We extend \(\mathbb{N}\) by adding the element \(\infty\), denoting infinity, forming the set of extended  natural numbers: % ML@KK: define macros for example for \mathbb{N}^\infty 
\( \mathbb{N}^\infty = \mathbb{N} \cup \{\infty\} \).
We define  the order relations for  $\mathbb{N}^\infty$ : $<_\infty$ and  $\leq_\infty$:

            \[
           \text{For } n,m \in \mathbb{N}^\infty, n <_\infty m \text{ if and only if }
            \begin{cases} 
                n \in \mathbb{N} \text{ and } m \in \mathbb{N} \text{ and } n < m,\\
                n \in \mathbb{N} \text{ and } m = \infty.
            \end{cases}
            \]
            \[
             \text{For } n,m \in \mathbb{N}^\infty,n \leq_\infty m \text{ if and only if }
            \begin{cases} 
                n \in \mathbb{N} \text{ and } m \in \mathbb{N} \text{ and } n \leq m,\\
                n \in \mathbb{N} \text{ and } m = \infty,\\
                n = \infty \text{ and } m = \infty.
            \end{cases}
            \]
\subsection{Intervals}


\begin{definition}[Interval]
A \emph{non-empty interval} is a pair $[\tmin,\tmax]$ with $\tmin \in \mathbb{N}$, $\tmax \in \mathbb{N}^\infty$, and $\tmin \leq_\infty \tmax$. The \emph{empty interval}, written $\varnothing$, contains no elements.
The set of all valid intervals is
%\[
defined as $
\mathbb{I} := \{[\tmin,\tmax] \mid \tmin \leq_\infty \tmax\} \cup \{\varnothing\}.
$
%\]
For $n\in\mathbb{N}$, we write $n\in[\tmin,\tmax]$ iff $\tmin \leq_\infty n \leq_\infty \tmax$. 
 \end{definition}

 Informally, a non-empty interval $[\tmin,\tmax]$  \emph{represents} a contiguous set of numbers, that is:  $\{n \mid n\in[\tmin,\tmax]\}$, and the empty interval $\varnothing$ represents the empty set of natural numbers.
The \emph{size} of an interval $I$, written $\size{I}$, relates to the cardinality of the set that it represents. Thus, the size for $\varnothing$ is zero and for valid non-empty intervals:
\[
\size{[a,b]} :=
\begin{cases}
b-a-1 & \text{if } b\!\in\!\mathbb{N},\\
\infty & \text{if } b=\infty.
\end{cases}
\]

  \begin{example} 
  Consider the following examples of intervals:
\begin{itemize}
    \item If \(\tmin := 2\) and \(\tmax := 5\), then \(I := [2, 5] \text{ represents } \{ 2, 3, 4, 5 \}\).
    \item If \(\tmin := 3\) and \(\tmax := \infty\), then \(I := [3, \infty] \text{ represents }\{ t \in \mathbb{N} \mid t \geq 3 \}\).
    \item If \(\tmin := 3\) and \(\tmax := 3\), then \(I := [3, 3] \text{ represents } \{ 3\}\).
     \item If \(\tmin := 3\) and \(\tmax := 0\), then $I:=[3,0]$ is not a valid interval.
\\

     We transform the timed constraints from the  use case to intervals:
     \begin{enumerate}
         \item Let $I_1:=[9,16]$ represent the timed constraint between 9:00 and 16:00 from the Delivery Clause (Table~\ref{tab:contract}).
         \item Let $I_2:=[8, 24]$ represent the constraint between 8:00 and 24 from the Maintenance Announcement (Table~\ref{tab:ma}).
         \item Let $I_3:=[10,14]$ represent the constraint between 10:00 and 14:00 from the Safety regulation (Table \ref{tab:contract}).
     \end{enumerate}
\end{itemize}
\end{example}


\subsection{Allen's Algebra and Unifiability}


Allen's algebra \cite{allen1983maintaining} consists of 13 exhaustive and mutually exclusive relations between \textit{abstract}\footnote{Not inherently tied to a specific numerical domain.} pairs of intervals. To provide an intuition, in Fig.~\ref{fig:ia} we give a visual interpretation of these relations. The different cases are based on the relative comparison of the start and end points of the intervals. In Table \ref{tab:IAdesc} we specify the constraints associated with each case for the domain of intervals $\mathbb{I}$ defined on natural numbers.


\begin{figure}[h]
    \begin{subfigure}{0.22\textwidth}
        \centering
          \begin{tikzpicture}[scale=0.5, yscale=0.80]
            \draw[->,gray] (-0.5,0) node[left] {} -- (3.75,0);
            \drawinterval{.5}{1.7}{2.5}{1.7}{$I$}
            \drawinterval{.5}{.7}{2.5}{.7}{$I'$}
            \node[] at (2,3) {$I$ equals $I'$ ($eq$)};
        \end{tikzpicture}
        %\caption*{$I$ equals $I'$ ($eq$)}
        \label{subfig:ia:eg}
    \end{subfigure}
    \hfill
    \begin{subfigure}{0.19\textwidth}
        \centering
          \begin{tikzpicture}[scale=0.5, yscale=0.80]
            \draw[->,gray] (-0.5,0) node[left] {} -- (4,0);
            \drawinterval{0}{1.7}{1.4}{1.7}{$I$}
            \drawinterval{1.9}{.7}{3.35}{.7}{$I'$}
            \node[] at (2,4) {$I$ before $I'$ ($b$)};
            \node[] at (2,3) {$I'$ after $I$ ($bi$)};
        \end{tikzpicture}
        \label{subfig:ia:b}
    \end{subfigure}
    \hfill 
    \begin{subfigure}{0.22\textwidth}
        \centering
          \begin{tikzpicture}[scale=0.5, yscale=0.80]
            \draw[->,gray] (-0.75,0) node[left] {} -- (3.75,0);
            \drawinterval{0}{1.7}{1.5}{1.7}{$I$}
            \drawinterval{1.5}{.7}{3}{.7}{$I'$}
            \node[] at (2,4) {$I$ meets $I'$ ($m$)};
            \node[] at (2,3) {$I'$ met by $I$ ($mi$)};
        \end{tikzpicture}
        \label{subfig:ia:m}
    \end{subfigure}
    \hfill 
    \begin{subfigure}{0.22\textwidth}
        \centering
          \begin{tikzpicture}[scale=0.5, yscale=0.80]
            \draw[->,gray] (-0.5,0) node[left] {} -- (3.75,0);
            \drawinterval{0.5}{1.7}{2.5}{1.7}{$I$}
            \drawinterval{0}{.7}{3.1}{.7}{$I'$}
            \node[] at (2,4) {$I$ during $I'$ ($d$)};
            \node[] at (2,3) {$I'$ contains $I$ ($di$)};
        \end{tikzpicture}
        \label{subfig:ia:d}
    \end{subfigure}
    \vspace{0.05cm}
    \quad
    \hfill 
    \begin{subfigure}{0.23\textwidth}
        %\caption*{$I$ starts $I'$ ($s$)\\ $I'$ started by $I$ ($si$)}
        \centering
          \begin{tikzpicture}[scale=0.5, yscale=0.80]
            \draw[->,gray] (-0.5,0) node[left] {} -- (3.75,0);
            \drawinterval{0}{1.7}{1.15}{1.7}{$I$}
            \drawinterval{0}{.7}{3.1}{.7}{$I'$}
            \node[] at (1.75,-1) {$I$ starts $I'$ ($s$)};
            \node[] at (1.75,-2) {$I'$ started by $I$ ($si$)};            
        \end{tikzpicture}
        \label{subfig:ia:s}
    \end{subfigure}
    \hfill
    \begin{subfigure}{0.22\textwidth}
        %\caption*{$I$ overlaps $I'$ ($o$)\\ $I'$ overlapped by $I$ ($oi$)}
        \centering
          \begin{tikzpicture}[scale=0.5, yscale=0.80]
            \draw[->,gray] (-0.75,0) node[left] {} -- (3.75,0);
            \drawinterval{0}{1.7}{1.5}{1.7}{$I$}
            \drawinterval{1.2}{.7}{2.8}{.7}{$I'$}
            \node[] at (1.5,-1) {$I$ overlaps $I'$ ($o$)};
            \node[] at (1.5,-2) {$I'$ overlapped by $I$ ($oi$)};
        \end{tikzpicture}
        \label{subfig:ia:o}
    \end{subfigure}
    \hfill
    \begin{subfigure}{0.25\textwidth}
        \centering
          \begin{tikzpicture}[scale=0.5, yscale=0.80]
            \draw[->,gray] (-0.5,0) node[left] {} -- (3.75,0);
            \drawinterval{2.1}{1.7}{3.1}{1.7}{$I$}
            \drawinterval{0}{.7}{3.1}{.7}{$I'$}
            \node[] at (1.5,-1) {$I$ finishes $I'$ ($f$)};
            \node[] at (1.5,-2) {$I'$ finished by $I$ ($fi$)};
        \end{tikzpicture}
        %\caption*{$I$ finishes $I'$ ($f$)\\ $I'$ finished by $I$ ($fi$)}
        \label{subfig:ia:f}
    \end{subfigure}
    \hfill
    \quad
    \hfill
    \caption{Visual representation of all base relations from Allen's Interval Algebra (continuous Intervals)}
    \label{fig:ia}
\end{figure}



 We additionally define the \emph{non-unifiable} relation $\overline{U}$: the relation captures that the union of two intervals is not closed under $\mathbb{I}$. This relation is a special case of the \textit{before} relation: $$\overline{U}([a,b],[c,d]) \text{ iff } ([a,b+1] \text{ before } [c,d]) \sor ([c,d+1] \text{ before } [a,b]) .$$


\begin{example}

We show below examples illustrating some relations between intervals:

\begin{enumerate}
    \item \textbf{Before}: \( I := [0, 3] \) and \( I' := [4, 7] \). \\
    \( I \text{ is before } I' \), since \( I \) ends before \( I' \) starts.

    \item \textbf{Non-unifiable}: \( I := [0,2] \) and \( I' := [4, 7] \). \\
    \( I \text{ is not unifiable with } I' \), since $[0,3]$ precedes $[4,7]$.
    
    \item \textbf{After}: \( I := [5, 8] \) and \( I' := [0, 4] \). \\
    \( I \text{ is after } I' \), since \( I \) starts after \( I' \) ends.
    
    \item \textbf{Meets}: \( I := [0, 3] \) and \( I' := [3, 6] \). \\
    \( I \text{ meets } I' \), since \( I \) ends exactly when \( I' \) starts.
    
    \item \textbf{Overlaps}: \( I := [0, 4] \) and \( I' := [3, 6] \). \\
    \( I \text{ overlaps } I' \), since \( I \) starts before \( I' \) and ends after \( I' \) starts.
    
    \item \textbf{Starts}: \( I := [2, 5] \) and \( I' := [2, 8] \). \\
    \( I \text{ starts } I' \), since \( I \) and \( I' \) have the same start time, but \( I \) ends before \( I' \).
    
    \item \textbf{Finishes}: \( I := [4, 7] \) and \( I' := [1, 7] \). \\
    \( I \text{ finishes } I' \), since \( I \) and \( I' \) have the same end time, but \( I \) starts after \( I' \).
    
    \item \textbf{During}: \( I := [3, 6] \) and \( I' := [1, 8] \). \\
    \( I \text{ is during } I' \), since \( I \) starts after \( I' \) starts and ends before \( I' \) ends.
    
    \item \textbf{Contains}: \( I := [0, 7] \) and \( I' := [2, 5] \). \\
    \( I \text{ contains } I' \), since \( I \) starts before \( I' \) and ends after \( I' \) ends.
    \item \textbf{Non-unifiable}: \( I := [0, 3] \) and \( I' := [5, \infty] \). \\
    \( \overline{U}(I,I') \), since $3+1 < 5$.
\end{enumerate}
\end{example}

\begin{table}[t]
    \centering
    \begin{tabular}{|l|l|l|}
    \hline
        Notation& Interpretation  & Necessary constraints  \\
        \hline
        $\boldsymbol{I< I'}$  & $I$ precedes $I'$  & $\max(I) <_\infty \min(I')$\\
        $\boldsymbol{I' > I}$  & $I'$ precedes $I$  &  \\
        \hline
        $\boldsymbol{I = I'}$  & $I$ is equal to $I'$  & $\min(I) = \min(I')$, $\max(I) = \max(I')$ \\
        & $I'$ is equal to $I$  & \\
        \hline
        $\boldsymbol{I~m~I'}$  & $I$ meets $I'$  & $\max(I) =\min(I')$ \\
        $\boldsymbol{I'~mi~I}$  & $I'$ meets $I$  &  \\
        \hline
        $\boldsymbol{I~o~I'}$  & $I$ overlaps with $I'$  & $\min(I) <_\infty \min(I') < \max(I) <_\infty \max(I')$ \\
        $\boldsymbol{I'~oi~I}$  & $I'$ overlaps with $I$  & \\
        \hline
        $\boldsymbol{I~s~I'}$  & $I$ starts $I'$  & $\min(I) = \min(I')$, $\max(I) <_\infty \max(I')$ \\
        $\boldsymbol{I' si I}$  & $I'$ starts $I$  &  \\
        \hline
        $\boldsymbol{I~d~I'}$  & $I$ during $I'$  & $\min(I') <_\infty \min(I)$, $\max(I) <_\infty \max(I')$ \\
        $\boldsymbol{I'~di~I}$  & $I'$ during $I$  & \\
        \hline
        $\boldsymbol{I~f~I'}$  & $I$ finishes $I'$  & $\max(I) = \max(I')$, $\min(I) >_\infty \min(I')$ \\
        $\boldsymbol{I'~fi~I}$  & $I'$ finishes $I$  &  \\
        \hline
    \end{tabular}
    \caption{Constraint on intervals from $\mathbb{I}$ to satisfy a relation from  Allen's Algebra }
    \label{tab:IAdesc}
\end{table}


The {intersection} operator: $ \cap:\mathbb{I} \times \mathbb{I}  \to \mathbb{I}$  is defined as follows:
	\begin{align*}
		[a,b] \cap [c,d] := 	\left\{
		\begin{array}{lll}
			~	\varnothing & \text{iff} &  [a,b] \text{ before } [c,d]\text{ or } [c,d] \text{ before } [a,b] \\
			~ [max(a, c), min(b,d)] & \text{otherwise}
		\end{array}
		\right.
	\end{align*}

The operators we define now are partially defined as the interval domain not being closed under these operations.

The \emph{union} operator $ \cup:\mathbb{I} \times \mathbb{I}  \rightharpoonup \mathbb{I}$ is partially defined as follows:

 $$[a,b] \cup [c,d] := \begin{cases} 		
		\text{ undefined }& \text{ iff } \overline{U}([a,b],[c,d]) ,\\
		[min(a,b,c,d),max(a,b,c,d)] & \text{otherwise.}
	\end{cases}$$

The partial interval difference operator $\ominus: \mathbb{I} \times \mathbb{I} \rightharpoonup \mathbb{I}$ is the non-commutative, partially defined operation removing all elements from the interval  $I'$ from the interval $I$ in $I \ominus I'$:

$$[a,b] \ominus [c,d] := \begin{cases} 		
		\text{ undefined }& \text{ iff   } [c,d] \text{ during } [a,b],\\
		[a,b] & \text{iff   } [a,b] \cap [c,d]= \emptyset,\\
  \varnothing & \text{iff   } [a,b] \subseteq [c,d],\\
  [a,c-1] & \text{iff   } [a,b] \text{ meets or overlaps or finished by } [c,d],\\
  [d+1,b] & \text{iff   } [c,d] \text{ meets or overlaps or starts by } [a,b].\\
	\end{cases}$$

\begin{example}

Let us consider some examples concerning the operator $\ominus$:

    \begin{itemize}
    \item $
    [2, 8] \ominus [4, 6] = \text{undefined} \quad \text{(during)}$
    
    \item $
    [0, 3] \ominus [5, 7] = [0, 3] \quad \text{(empty intersection)}$
    \item $
    [2, 5] \ominus [1, 6] = \varnothing \quad \text{(subset)}
    $  
    \item $[1, 5] \ominus [3, 7] = [1, 2] \quad \text{(overlaps on left)}$
    \item $[0, 5] \ominus [3, \infty] = [0, 2] \quad \text{(overlaps left with infinity)}$
    \item $[4, 8] \ominus [1, 5] = [6, 8] \quad \text{(overlaps on right)}$
    \item $[3, \infty] \ominus [0, 5] = [6, \infty] \quad \text{(overlaps right with infinity)}$
\end{itemize}


We use now the difference operator on the three-time intervals from our use case:
$I_1 \ominus I_2 = [9,16]\ominus [12,\infty]= [9,11]$. 
On the other hand,
$I_1 \ominus I_3 = [9,16] \ominus [10,14]$ is undefined as $[10,14]$ during $[9,16]$.

\end{example}

\subsection{Interval Sets}
As demonstrated in the previous subsection, the interval domain $\mathbb{I}$ is not closed under union and difference operations. To address this limitation, we introduce the concept of interval sets.
An \emph{interval set}, denoted as $\is$, is a finite set of intervals equipped with a condition to ensure minimal representation.
This condition is called the \emph{canonical form}, which ensures that the interval set contains the smallest number of non-overlapping and non-adjacent intervals necessary to represent a given set of time points.


\begin{definition}[Canonical Interval Set]\label{def:canonis}
An \defn{interval set} $\is$ is said to be in \emph{canonical form} if no two intervals in $\is$ are unifiable. Formally:
\[
\forall \{I, I'\} \subseteq \is: \overline{U}(I, I').
\]
where $\overline{U}(I, I')$ denotes that $I$ and $I'$ are not unifiable (i.e., $I \cup I'$ is not an interval from $\mathbb{I}$).

For the remainder of this paper, we adopt the following notational convention: we use $\is^*$ to refer to interval sets composed exclusively of finite (bounded) intervals. In contrast, we use $\is$ to denote interval sets that may contain infinite intervals, such as those extending to $+\infty$. 

The \emph{size of a set of intervals} $\is$, written $\size{\is}$, is defined as the sum of the sizes of the individual intervals it contains:
\[
\size{\is} := \sum_{I \in \is} \size{I}.
\]
The \emph{cardinality} of the $\is:=\{I_0,I_1, \dots,I_n\}$, written $\card(\is)$, is the number of element intervals in the set,
$\size{\is} := n+1$.
\end{definition}




\begin{example}[Canonical form for interval sets]
Consider the interval set $\is':= \{ [0, 2], [3, 6] \}$. The two intervals forming this interval set are unifiable as $[0, 2] \cup [3, 6] = [0,6]$. The canonical representation is thus $\is := \{ [0, 6] \}$.

Consider the interval set $\is:= \{ [0, 2], [5, 7], [10, \infty] \}$.\\ The pair $[0, 2]$ and $[5, 7]$ satisfy $\overline{U}([0, 2], [5, 7])$.\\ Similarly, $[5, 7]$ and $[10, \infty]$ satisfy $\overline{U}([5, 7], [10, \infty])$.\\ Finally, $[0, 2]$ and $[10, \infty]$ also satisfy $\overline{U}([0, 2], [10, \infty])$.
\end{example}

These representations enable a symbolic encoding of timed constraints, such as those described in our use case. For instance, the obligations from \MA, as shown in Table~\ref{tab:contract}, that an action is permitted \textit{between 02:00 and 04:00, and between 08:00 and 24:00}—can be expressed as the canonical interval set:
\[
\is_3 := \{ [2,4], [8,24] \}.
\]
Similarly, constraints from \DC and \SR can each be represented using singleton interval sets. The constraint \textit{between 09:00 and 16:00}, for example, corresponds to:
\[
\is_1 := \{ [9,16] \},
\]
while \textit{between 10:00 and 14:00} is represented as:
\[
\is_2 := \{ [10,14] \}.
\]
We adopt this interval-set representation as a uniform basis for the symbolic manipulation of temporal constraints within the logical framework developed in the following sections.

\begin{definition}[Expansion Factor of an Interval Set]
\label{def:expansion-factor}
Let \(\is\) be a finite set of intervals. The \defn{expansion factor} \(\expf(\is)\) of \(\is\) is defined as the ratio between the total number of discrete points covered by the intervals in \(\is\) and the number of intervals in \(\is\), i.e.,
\[
\expf(\is) := \frac{\size{\is}}{\card(\is)}\quad,
\]
where \(\size{\is}\) is the sum of the sizes (number of points) of the intervals in \(\is\), and \(\card(\is)\) is the number of intervals in \(\is\).
\end{definition}

\begin{example}[Minimal Expansion Factor]
Consider the interval set
\[
\is_1 = \{ [3,3], [5,5], [10,10] \}.
\]
The cardinality is $\card(\is_1) = 3$, and the size is
\[
\size{\is_1} = (3 - 3 + 1) + (5 - 5 + 1) + (10 - 10 + 1) = 1 + 1 + 1 = 3.
\]
Therefore, the expansion factor is
\[
\expf(\is_1) = \frac{3}{3} = 1,
\]

\end{example}

\begin{example}[Large Expansion Factor]
Consider now the interval set
\[
\is_2 = \{ [1,1000], [1050,1050], [2000,2000] \},
\]
which consists of one large interval and two punctual intervals. Here,
\[
\card(\is_2) = 3,
\]
and
\[
\size{\is_2} = (1000 - 1 + 1) + (1050 - 1050 + 1) + (2000 - 2000 + 1) = 1000 + 1 + 1 = 1002.
\]
Thus, the expansion factor is
\[
\expf(\is_2) = \frac{1002}{3} = 334,
\]
Showing a large blow-up, mainly caused by the large intervals.
\end{example}

\subsection{Arithmetic of Interval Sets}\label{isarithmetic}
    \paragraph{Sub-element} An interval $I$ is a \emph{sub-element} of $\mathfrak{I}$ (written $I \subin \mathfrak{I}$) if it is included in one of the elements of the set of intervals. We write:
    $ I \subin \mathfrak{I} \text{ iff } \exists I' \in \mathfrak{I} : I \subseteq I' $. We abuse the notation for time points $t  \in \mathbb{N}$ and write $t \subin \mathfrak{I}$ iff $[t,t] \subin \mathfrak{I}$.

    \paragraph{Inclusion} We say $\mathfrak{I}$ is \emph{included} in $\mathfrak{I}'$ and write:
     $\mathfrak{I} \subset \mathfrak{I}' \text{ iff } \forall I \in \mathfrak{I} ~\exists I' \in \mathfrak{I}': I \subset I'$.

    \begin{example} Given $\mathfrak{I}:= \{ [1, 4], [6, 9] \}$:
    \begin{itemize}
        \item The interval $[2, 3] \subin \mathfrak{I}$ because $[2, 3] \subseteq [1, 4] \in \mathfrak{I}$.
        \item The time point $7 \subin \mathfrak{I}$ because $[7, 7] \subseteq [6, 9] \in \mathfrak{I}$.
    \end{itemize}
%    \end{example}

    Let $\mathfrak{I'} = \{ [0, 4], [6, \infty] \}$. Then $\mathfrak{I} \subseteq \mathfrak{I'}$ since
    \begin{itemize}
        \item $[1, 4] \subseteq [0, 4] \in \mathfrak{I'}$,
        \item $[6, 9] \subseteq [6, \infty] \in \mathfrak{I'}$.
    \end{itemize}
    
    For the sets of intervals $\is_1:= \{[9,16]\}$ and $\is_3:=\{[2,4],[8,24]\}$ from 
    our use case we have: $\is_1 \subset \is_3$.
    \end{example}

    \paragraph{Intersection} The intersection of two sets of intervals $\mathfrak{I}$ and $\mathfrak{I}'$ is the set of intervals containing all the intervals present in both interval sets. The operator is commutative and is defined as:
    $$
    \mathfrak{I} \cap \mathfrak{I}' :=\begin{cases}
     \emptyset& \text{if } \forall I\in \mathfrak{I}, \forall I'\in \mathfrak{I}': I \cap I'=\varnothing,\\
     \{ I \cap I' \mid I \in \mathfrak{I}, I' \in \mathfrak{I}', I \cap I' \neq \varnothing \} & \text{otherwise}.
    \end{cases}
    $$

    \begin{example} Consider the following examples:
    \begin{itemize}
        \item If $\mathfrak{I}:= \{ [1, 5], [8, 10], [19,\infty] \}$ and $\mathfrak{I}':= \{ [3, 7], [9, 12] \}$, we have $\mathfrak{I}\cap \mathfrak{I}' = \{ [3, 5], [9, 10] \}$,\\ as $[1, 5] \cap [3, 7] = [3, 5]$ and $[8, 10] \cap [9, 12] = [9, 10]$.
        \item If $\mathfrak{I} := \{ [1, 5], [8, 10] \}$ and $\mathfrak{I}':= \{ [6, 7], [12, \infty] \}$ then $\mathfrak{I}\cap \mathfrak{I}' = \emptyset$ (as none of the intervals from $\mathfrak{I}$ intersect with any interval from $\mathfrak{I}'$).
    \end{itemize}

    For the intervals $\is_1$, $\is_2$, and $\is_3$ related respectively to $\DC$, $\SR$, and $\MA$ from 
    the use case:
    \begin{itemize}
        \item $\is_1 \cap \is_2= \{[9,16]\} \cap \{[10,14]\}= \{[10,14]\}$.
        \item $\is_1 \cap \is_3 = \{[9,16]\} \cap \{[2,4],[8, 24]\}= \{[9,16]\}$.
    \end{itemize}
\end{example}


 \subsubsection{Union}
The union of two interval sets $\is$ and $\is'$ is recursively computed by the successive union of each interval $I_i \in \is'$ with $\is$:

\[
\is \cup \is' := \left\{ \left( \left(\is \cup I_1) \cdots \right) \cup I_n \right) \ \middle|\ I_1, \dots, I_n \in \is' \right\}.
\]


The union of an interval set $\is$ with a single interval $I$ is given by:
\[
\is \cup I := \left\{ \left( \bigcup_{I_i \in U(I, \is)} (I \cup I_i) \right) \right\} \circ \overline{U}(I, \is),
\]

where $\circ$ denotes set concatenation, and the subset of unifiable and non unifiable interval with $I$ from $\is$ are defined as:
\begin{align*}
\overline{U}(I, \is) &:= \{ I_i \in \is \mid \overline{U}(I, I_i) \}, \\
U(I, \is) &:= \{ I_i \in \is \mid I_i \notin \overline{U}(I, \is) \}.
\end{align*}

\begin{example} Let \(\is := \{ [1, 5], [9, 11] \}\) and \(\is' := \{ [3, 7], [9, 12] \}\):
\begin{align*}
\is \cup \is' &= \big(\{ [1, 5], [9, 11] \}\cup [1,7]\big) \cup [9,12] &\text(unrolling)\\
&= \big( \{ [1,5] \cup [3,7]\} \circ\{ [9,11]\}\big) \cup [9,12]  & (\text{as~}  \overline{U}([9,11], [3,7]))\\
&=\{[1,7],[9,11]\} \cup [9,12] & \text{(simplification)}\\
&= \{ [9,11] \cup [9,12] \} \circ \{[1,7]\}  & (\text{as~}\overline{U}([9,12], [1,7]))\\
&= \{[1,7],[9,12]\}
\end{align*}
\end{example}

\subsubsection{Difference (Reduction)}\label{differenceop}

The \emph{difference} of an interval set $\is$ by an interval set $\is'$, denoted $\is \ominus \is'$, is the set of intervals in $\is$ excluding those that overlap with $\is'$:

$$\is \ominus \is' := \bigcup_{I_i \in \is}(I \ominus \is') \text{ with } I \ominus \mathfrak{I'}= \bigcap_{I_i \in \is'}(I \ominus I_i).$$
Where the \emph{redefined} reduction of two intervals is now computed as \emph{set of intervals} instead of intervals as defined in the arithmetic of intervals: $\ominus: \mathbb{I} \times \mathbb{I} \to 2^\mathbb{I}$ is defined as :
	$$[a,b] \ominus [c,d] := \begin{cases} 		
		\{[a,c-1], [d+1,b]\}& \text{ iff   } [c,d] \text{ during } [a,b],\\
		\{[a,b]\} & \text{iff   } [a,b] \cap [c,d]= \varnothing,\\
  \emptyset & \text{iff   } [a,b] \subseteq [c,d],\\
  \{[a,c-1]\} & \text{iff   } [a,b] \text{ meets or overlaps or finished by} [c,d],\\
  \{[d+1,b]\} & \text{iff   } [c,d] \text{ meets or overlaps or starts by} [a,b].\\
	\end{cases}$$
    
\begin{remark}
An alternative definition could be $I \ominus I' := \{I_i \mid I_i \subseteq I \text{ and } I_i \not\subseteq I'\}.$ 
\end{remark}

\begin{example} Let $\is = \{ [1, 10] \}$ and $\is' = \{ [3, 5],\ [7, 8] \}$:

\begin{itemize}
    \item $[1, 10] \ominus [3, 5]= \{[1,2], [6,10]\}$.
    \item $[1, 10] \ominus [7, 8]= \{[1,6], [9,10]\}$.
\end{itemize}
\end{example}

Thus, $\is \ominus \is' = \{ [1, 2],\ [6, 6],\ [9, 10] \}$.

From the intervals $\is_1, \is_2, \is_3$ related to our three normative systems from the use case, we have:
\begin{itemize}
    \item $\is_1 \ominus \is_2= \{[9,16]\} \ominus \{[10,14]\}= \{[9,9],[15,16]\}.$
    \item $\is_1 \ominus \is_3 = \{[9,16]\} \ominus \{[2,4],[8, 24]\}= \emptyset. $
\end{itemize}

\input{traces.tex}