\documentclass[11pt,a4paper]{article}

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{geometry}
\usepackage{enumitem}
\usepackage{xcolor}
\usepackage{hyperref}
\usepackage{longtable}
\usepackage{array}

\geometry{margin=2.5cm}
\hypersetup{
  colorlinks=true,
  linkcolor=blue,
  urlcolor=blue
}
\setlist[itemize]{leftmargin=*, itemsep=0.35em}
\setlist[enumerate]{leftmargin=*, itemsep=0.35em}
\newcommand{\sev}[1]{\textbf{#1}}
\newcommand{\loc}[1]{\texttt{#1}}

\title{Final Dissertation Review Report}
\author{Generated review for \texttt{thesis.tex}}
\date{12 July 2026}

\begin{document}
\maketitle

\section*{Scope and Method}

This report reviews the active dependency tree of \loc{thesis.tex}. The review
covered the main file, front matter, chapters, transitive inputs, style and macro
files, and the current build logs. No existing thesis source file was changed.

The review used four read-only passes:
\begin{itemize}
  \item \textbf{LaTeX/build/style checker:} build-log warnings, labels, references,
  theorem setup, PDF metadata, overfull boxes, and source-submission debris.
  \item \textbf{Fact and citation checker:} internal consistency, scenario logic,
  terminology, citation coverage, and bibliography warnings using local files.
  \item \textbf{Readability checker:} argument flow, unclear transitions, overly dense
  sections, informal conclusion language, and inconsistent scope framing.
  \item \textbf{Spelling, grammar, and prose-risk checker:} typos, grammar problems,
  repeated words, unnatural phrasing, and generic prose. This is not an AI-authorship
  detector; it only flags passages that may read as generic, vague, or synthetic.
\end{itemize}

\section*{Executive Summary}

\begin{itemize}
  \item \sev{Critical before submission:} fix the wrong PDF metadata in
  \loc{thesis.tex:8}; fix math-mode Unicode/corruption in
  \loc{forwardtight1.tex:406--411}; fix the contradictory blame assignment in
  \loc{chaptermonitor.tex:242}; fix the duplicated sentence in \loc{fazit.tex:36}.
  \item \sev{Major consistency issue:} the dissertation scope says timed and
  distributed normative systems, but the two technical parts remain split:
  metric-time single-agent conflict analysis and logical-time two-agent monitoring.
  Add an explicit bridge sentence in the introduction and conclusion.
  \item \sev{Major prose issue:} the conclusion currently has several typos and
  informal phrases. It should be edited first because it is short, visible, and
  central to the final impression.
  \item \sev{Build status from existing logs:} no active duplicate labels were found,
  and the current \loc{thesis.log} shows no undefined references or citations.
  However, it does show theorem-style warnings, invalid text commands in math mode,
  missing glyphs, and significant overfull boxes.
\end{itemize}

\section*{Priority 1: Must Fix Before Submission}

\begin{enumerate}
  \item \sev{Wrong PDF title metadata.} \loc{thesis.tex:8} still sets
  \verb|pdftitle={Wie schreibe ich eine Masterarbeit?}|. Submission systems and
  library repositories often index PDF metadata. Replace it with the actual
  dissertation title from \loc{title.tex:26}.

  \item \sev{Corrupted math-mode text.} \loc{forwardtight1.tex:406--411} contains
  forms such as \verb|\tsmc©| and curly apostrophes in math, e.g. \verb|C’|.
  The log reports invalid \verb|\textcopyright| and \verb|\textquoteright| in math
  mode plus missing glyphs. Replace with standard math/ASCII syntax, likely
  \verb|\tsmc(C)| and \verb|C'|.

  \item \sev{Contradictory blame assignment.} In \loc{chaptermonitor.tex:242}, the
  tenant stops paying rent during the three-month notice period, but the verdict
  assigns blame to the landlord. This conflicts with C5 in
  \loc{chaptermonitor.tex:158--164}, which says the tenant must continue paying
  during the notice period. Either assign blame to the tenant or change the scenario
  so that the landlord obstructs payment or occupancy.

  \item \sev{Guarded termination examples conflict with C5.}
  \loc{quantitativeviol.tex:349--355} and \loc{qviol.tex:354--360} use
  \verb|\guard[\Gamma^+ \cdot \notifterm^{(1)}]{\repit{C_3}}| and discharge the
  contract immediately after notice. Later, \loc{forwardtight.tex:441--449} uses
  a three-step continuation with \verb|\Gamma^3|. Either add the continuation to
  these examples or state clearly that they are simplified examples, not C5.

  \item \sev{Garbled future-work sentence.} \loc{fazit.tex:36} contains the merged
  phrase ``dutiesallow regulators''. Rewrite as one clean sentence about verifying
  both feasible time windows and absence of forced interference between agents.

  \item \sev{Broken framework-workflow paragraph.} \loc{chapterconflict.tex:1497--1501}
  mixes SMT checking, DPNF, conflict density, calculus, repair, and interactive
  workflow in one dense block. It also contains the fragment ``Unless all the other
  tools in the framework are.'' Rewrite this subsection as a numbered workflow:
  input specification; SMT feasibility check; DPNF/conflict extraction; density;
  automatic pruning; optional calculus-guided refinement.
\end{enumerate}

\section*{Priority 2: Major Conceptual and Structural Edits}

\begin{enumerate}
  \item \sev{Clarify the thesis scope split.} \loc{einleitung.tex:135} frames the
  dissertation as timed and distributed ought-to-do normative systems. The actual
  contributions split into metric-time single-agent static conflict analysis and
  two-agent logical-time monitoring. Add a sentence explaining that the two parts
  deliberately isolate complementary dimensions, and that their integration is
  future work.

  \item \sev{Align chapter title with actual arity.} \loc{chaptermonitor.tex:1}
  says ``Multi-Party Collaborative Contracts'', but the framework is consistently
  two-agent/two-party. Either rename the chapter to ``Two-Party'' or add a clear
  scope note that two-party contracts are the base case for multi-party work.

  \item \sev{Resolve acronym/name inconsistency.} The German abstract
  \loc{abstract.tex:22} uses ``Two-Agents Collaborative Normative Logic (TACNL)'',
  while the English abstract \loc{abstract.tex:48} and chapters use \verb|\cDL|.
  Use one public name/acronym consistently across abstracts, chapter text, and
  conclusion.

  \item \sev{Strengthen related-work structure in Chapter 4.}
  \loc{qviol.tex:1491--1586} mixes closest formalisms, broad field, causality, and
  monitoring methodology. Normalize the hierarchy so each promised comparison
  dimension has a clear subsection.

  \item \sev{Check the quantitative blame lemma framing.} \loc{qviol.tex:868--880}
  asks whether blame counters ``always give back the same score'', but the lemma
  proves only \verb|\qsem{\pi,C} \le n_1+n_2|. Revise the question to match the
  inequality or strengthen the theorem if equality is intended.

  \item \sev{Add citation near SCM claims.} \loc{qviol.tex:1577--1580} discusses
  structural causal models without a nearby citation. The local bibliography
  contains \loc{halpern2014structural} and \loc{halpern2016causality}; cite the
  relevant one(s) near the SCM comparison.
\end{enumerate}

\section*{Priority 3: LaTeX and Final-Submission Hygiene}

\begin{enumerate}
  \item \sev{Unknown theorem styles.} \loc{style.tex:222} and \loc{style.tex:226}
  use \verb|\theoremstyle{thmstyletwo}| and \verb|\theoremstyle{thmstylethree}|.
  The log reports unknown theorem-style warnings for examples, remarks,
  definitions, lemmas, claims, and corollaries. Define these styles or use standard
  \verb|plain|, \verb|definition|, and \verb|remark|.

  \item \sev{Bookmark-safe math macros.} Section/bookmark titles such as
  \loc{chapterconflict.tex:373}, \loc{chapterconflict.tex:393},
  \loc{chapterconflict.tex:516}, \loc{chapterconflict.tex:603},
  \loc{chapterconflict.tex:932}, and \loc{conflictmeat.tex:922} contain
  \verb|\TDDLfm|. The log shows \loc{hyperref} PDF-string warnings. Use
  \verb|\texorpdfstring| or define a bookmark-safe version of the macro.

  \item \sev{Large overfull boxes.} The existing log reports visible overfull boxes,
  including around \loc{forwardtight1.tex:168} and \loc{forwardtight1.tex:417}.
  Split long inline constructions into aligned displays.

  \item \sev{Todo package and source debris.} \loc{style.tex:184} and
  \loc{macros.tex:38} both load \loc{todonotes}. No active todo was found in the
  compiled text, but final submission should load it once, preferably disabled.
  Commented reviewer TODO blocks remain in \loc{chapterconflict.tex:383--460};
  remove them before source archival.

  \item \sev{Bibliography warning.} \loc{thesis.blg:16} reports an invalid ISBN for
  the entry \loc{10.1145/3597503.3639093} in \loc{thesisz.bib:5}. Validate the ACM
  ISBN, add expected hyphenation if appropriate, or remove the ISBN field.
\end{enumerate}

\section*{Priority 4: Spelling, Grammar, and Local Prose Fixes}

\begin{longtable}{>{\raggedright\arraybackslash}p{0.25\linewidth}
                  >{\raggedright\arraybackslash}p{0.31\linewidth}
                  >{\raggedright\arraybackslash}p{0.34\linewidth}}
\textbf{Location} & \textbf{Issue} & \textbf{Suggested correction} \\
\hline
\loc{acknowledgment.tex:11} & Duplicate word and spelling: ``defense and and'', ``straight forward''. & ``defense and made ... straightforward''. \\
\loc{acknowledgment.tex:17} & Sentence fragment beginning ``While I regret not having worked...''. & Join with the next sentence or rewrite. \\
\loc{acknowledgment.tex:19} & ``inconditional'' and unclear antecedent for ``who''. & ``unconditional''; recast the sentence. \\
\loc{declaration.tex:13} & ``up to their last versions'' is unnatural. & ``latest versions available as of January 2026''. \\
\loc{einleitung.tex:95} & ``Model checking'' capitalized mid-sentence. & ``model checking''. \\
\loc{traces.tex:87} & Mazurkiewicz comparison ends abruptly. & Split the comparison from the design choice. \\
\loc{traces.tex:266} & ``is between logical-time words and metric-time words''. & ``lies between'' or ``occupies a middle ground between''. \\
\loc{chaptermonitor.tex:3} & Tense shift and gendered phrasing: ``We built'', ``his attempts''. & Use consistent present tense and ``their attempts'' or ``the agent's attempts''. \\
\loc{chaptermonitor.tex:7} & Unnatural phrase ``moves to sharper checking''. & ``checks compliance more precisely''. \\
\loc{chaptermonitor.tex:344--345} & Repetition and missing preposition. & ``period-based action-set traces'' and ``merge both agents' traces into collaborative objectives''. \\
\loc{chapterconflict.tex:937} & ``inexistance''. & ``nonexistence''. \\
\loc{chapterconflict.tex:1488} & ``exlosion''. & ``explosion''. \\
\loc{chapterconflict.tex:1570} & ``achievement achievement'' and unclear sentence. & Remove duplicate and rewrite the contrast. \\
\loc{chapterconflict.tex:1607} & ``aby nd'' and unclear relative clause. & ``and by''; clarify the sentence. \\
\loc{conflictmeat.tex:209} & ``deontic operator an action'', ``espcially''. & ``deontic operator and action'', ``especially''. \\
\loc{conflictmeat.tex:866} & ``a an'', ``less then the the''. & ``an'', ``less than the''. \\
\loc{tightmotivation.tex:1067} & ``the the'' and stray final ``.x''. & Remove duplicate ``the'' and stray ``x''. \\
\loc{quantitativeviol.tex:9} & ``did every agent contributed''. & ``did each agent contribute''. \\
\loc{quantitativeviol.tex:478} & ``recusive'' and repeated ``application''. & ``recursive applications of the progression function''. \\
\loc{forwardtight.tex:17} & Sentence fragment after alphabet phrase. & Combine with the following sentence. \\
\loc{forwardtight.tex:343} & ``n repition''. & ``the n-th repetition''. \\
\loc{forwardtight.tex:488} & ``furthers'', ``definion'', awkward ``disjointedness''. & Rewrite the paragraph. \\
\loc{forwardtight1.tex:333--335} & ``unsatifiability'', ``raises whe'', ``easly spotable'', ``satifaction''. & Correct spelling and combine sentence fragments. \\
\loc{forwardtight2.tex:198} & ``they is responsible''. & ``agent 2 is responsible''. \\
\loc{forwardtight2.tex:282} & ``is constructed is constructed''. & Remove repetition. \\
\loc{forwardtight2.tex:661--662} & ``memories blames'', ``responsability'', ``developp''. & ``remember blame'', ``responsibility'', ``develop''. \\
\loc{qviol.tex:1487} & ``payement''. & ``payment''. \\
\loc{qviol.tex:1504} & Fragment after ``Although \verb|\logic{CSL}|...''. & ``Although ..., \verb|\cDL| and \verb|\logic{CSL}| differ in that...''. \\
\loc{fazit.tex:3} & ``Future Works''. & ``Future Work''. \\
\loc{fazit.tex:10} & ``developped'', ``normalform'', informal ``spot'' and ``made a way''. & ``developed'', ``normal form'', ``identify'', ``defined a measure''. \\
\loc{fazit.tex:14} & ``allowing to explain each failure''. & ``allowing the framework to explain each failure''. \\
\loc{fazit.tex:25} & ``was focused'', ``was on a single agent setting'', ``two agent setting''. & ``focuses on'', ``single-agent setting'', ``two-agent setting''. \\
\end{longtable}

\section*{Readability and Argument-Flow Comments}

\begin{itemize}
  \item \loc{chaptermonitor.tex:11}: the overview mentions ``four semantics'' but
  does not name them. Name the layers immediately or defer the phrase until the
  methodology paragraph.
  \item \loc{chaptermonitor.tex:67}: ``both agents suggest them'' is informal and
  underdefined. Use the formal term used later, such as attempt, enable, or offer.
  \item \loc{tightmotivation.tex:422--435}: the transition from contracts to
  regular-language tightness is abrupt and partly repetitive. Add one bridge
  sentence explaining why contract evaluation is reduced to prefix classification.
  \item \loc{conflictmeat.tex:920--932}: the MUS analogy is dense. Add a small
  mapping: propositional clause set to DPNF product term; clause/literal role to
  punctual literal; MUS to punctual conflict inside one product term.
  \item \loc{chapterconflict.tex:1565--1577}: the related-work comparison with
  Colombo/Governatori contains repetition and unclear wording. Rewrite as a direct
  contrast between their state/assumption model and your internalized action
  feasibility assumption.
\end{itemize}

\section*{AI-Style and Generic-Prose Risk}

This section does not claim to detect AI authorship. It flags passages that may
read as generic, informal, or insufficiently dissertation-specific.
\begin{itemize}
  \item \loc{fazit.tex:5--10}: phrases such as ``work well'', ``keeping things
  simple'', and ``tough reasoning tasks'' are too informal for a final conclusion.
  Replace with precise terms such as controlled expressiveness, tractable reasoning,
  conflict detection, and blame attribution.
  \item \loc{chaptermonitor.tex:7}: ``moves to sharper checking'' sounds generic.
  Replace with a concrete description of what is checked: attempts, successful
  outcomes, interference, and responsibility.
  \item \loc{einleitung.tex:37--43}: several broad claims about compliance burden
  and digitalization would benefit from either a citation or more precise framing.
  Avoid sweeping claims without source support.
\end{itemize}

\section*{Final Editing Order}

\begin{enumerate}
  \item Fix the four must-fix consistency/build issues: metadata, corrupted math
  characters, C5/blame contradiction, and guarded-termination inconsistency.
  \item Edit \loc{fazit.tex} completely. It is short and currently contains several
  visible typos, informal phrases, and one garbled sentence.
  \item Clean the Chapter 3 framework workflow around \loc{chapterconflict.tex:1497}
  and the Chapter 4 related-work hierarchy around \loc{qviol.tex:1491}.
  \item Resolve LaTeX warnings: theorem styles, bookmark-safe macros, large overfull
  boxes, and invalid ISBN.
  \item Rebuild the thesis from a clean state and inspect \loc{thesis.log} for
  remaining warnings before submission.
\end{enumerate}

\end{document}
